Diffusion MRS tracks distinct trajectories of neuronal development in the cerebellum and thalamus of rat neonates

Version 1

Abstract

Abstract

It is currently impossible to non-invasively assess cerebellar cell structure during early development. Here we propose a novel approach to non-invasively and longitudinally track cell-specific development using diffusion-weighted magnetic resonance spectroscopy in combination with microstructural modelling. Tracking metabolite diffusion allows us to probe cell-specific developmental trajectories in the cerebellum and thalamus of healthy rat neonates from post-natal day (P) 5 to P30. Additionally, by comparing different analytical and biophysical microstructural models we can follow the differential contribution of cell bodies and neurites during development. The thalamus serves as a control region to assess the sensitivity of our method to microstructural differences between the regions. We found significant differences between cerebellar and thalamic metabolites diffusion properties. For most metabolites, the signal attenuation is stronger in the thalamus, suggesting less restricted diffusion compared to the cerebellum. There is also a trend for lower signal attenuation and lower ADCs with increasing age, suggesting increasing restriction of metabolite diffusion. This is particularly striking for taurine in the thalamus. We use biophysical modelling to interpret these differences. We report a decreased sphere fraction (or an increased neurite fraction) with age for taurine and total creatine in the cerebellum, marking dendritic growth. Surprisingly, we also report a U-shape trend for segment length (the distance between two embranchments in a dendritic tree) in the cerebellum agreeing with age-matching morphometry of openly available 3D-Purkinje reconstructions. Results demonstrate that diffusion-weighted MRS probes early cerebellar neuronal development non-invasively.

Introduction

The cerebellum is a well-conserved structure across species and particularly across mammals. The cerebellar cortex is a highly folded structure organized in three layers: the molecular layer (containing the notable Purkinje cell dendritic trees), the Purkinje layer (containing the somas of Purkinje cells), and the granular layer (containing the granule cells). Projections to the deep cerebellar nuclei form the cerebellar white matter. Despite its small size (compared to the cerebrum), it contains more than half of the total brain neurons: this is mainly due to the granule cells, very small and densely packed cells, some of the smallest of the brain, presenting a very simple morphology1. On the other hand, the Purkinje cells are one of the largest types of neurons, and most complex cells in the brain. Cerebellar development is protracted2, starting from birth (or the last trimester of pregnancy) to a few weeks old in rodents (or 2 y.o. in humans)3. The cerebellum is involved in many basic and high-level functions, from motor control to social skills, and it is known to be affected in a number of neurodevelopmental disorders4.

During early development, neurons arborise. Their morphology becomes more complex and exhibits denser branching. In rodents, Purkinje cells’ dendritic trees start growing at birth. In 30 days, they evolve from a simple soma (∼10 µm diameter) to a large cell with a highly branched and planar dendritic tree (soma diameter ∼20 µm, area ∼10,000 µm2 in the mouse5). Purkinje cell morphology can be altered in autism spectrum disorders (ASD), as shown, post-mortem, in humans6, and in multiple mouse models of ASD. Specifically, they show an increased spine density in Tsc1 mutant mice7, an overgrowth of the dendritic tree and abnormal branching in PTEN KO mice8, a reduction of the dendritic tree in AUTS2 mice9; and general alterations in developmental disorders10–13. While cerebellar cell morphology is informative about brain development, there are very few translational techniques that allow for a non-invasive measure of their morphology.

All cerebellar outputs transit via the deep cerebellar nuclei (DCN), and a couple of them project to the thalamus before spreading to the cortex. The thalamus is considered as a relay station in the brain and contains about 60 nuclei on unique pathways. Some thalamic nuclei are involved in psychiatric disorders, and more specifically, the microstructure of the mediodorsal nucleus and the pulvinar nucleus are altered in schizophrenia and psychosis14. Some other nuclei are involved in neurodevelopmental disorders (e.g., the lateral reticular nuclei in ASD15). Generally, these disorders are associated with an altered thalamic connectivity14. Thalamocortical projections are crucial for the cytoarchitectural development of the cortex16,17.

Rodents reach adulthood after a couple of months and are an ideal model to study mammal brain development. Manganese-enhanced MRI (MEMRI) is particularly suited to study whole-brain early development in rodents, providing high structural contrast and speed of acquisition in neonates2,18, but the method is blind to the nature of the microstructural changes leading to the volumetric changes reported.

Magnetic resonance imaging (MRI) is the gold standard tool to access the properties of soft tissues in the body in a safe and non-invasive manner. Diffusion-weighted MRI (dMRI) measures the diffusion properties of water in the brain: constrained and hindered by cell membranes, water diffusion provides an indirect measure of the tissue microstructure19. However, water is ubiquitous, and can only partially disentangle the contribution from different tissue compartments (e.g. intra and extracellular space). Diffusion-weighted magnetic resonance spectroscopy (dMRS) measures the diffusion properties of metabolites in a large volume of interest. Brain metabolites are mostly intracellular, with some more specific to neurons (e.g., N-acetyl-aspartate, NAA, and glutamate, Glu) and others more specific to glial cells (e.g., choline compounds, tCho, and myo-inositol, Ins), therefore different cellular compartments can be resolved20. However, metabolites are 104-fold less concentrated than water, which necessitates large measurement volumes for dMRS (typically ∼8cm3 vs ∼8mm3 in dMRI in humans). Morphometric parameters can be extracted from dMRS experiments with multiple diffusion times and multiple diffusion-weightings, and they correlate with neuronal and glial morphologies extracted from histological measures in the healthy, adult mouse21,22. More specifically, it has been shown that myo-inositol diffusion properties reflect astrocytic morphology in an induced model of astrocytic hypertrophy23, and in a mouse model of demyelination where astrocytic activation was expected24. Choline compound’s diffusion properties were associated with microglial activation in human disease25, or in an LPS-induced brain inflammation model26. However, to the best of our knowledge, very few studies have used dMRS to probe changes in neuronal morphology, and none have attempted to probe neuronal development in early life.

Here, we propose a non-invasive method to monitor cellular brain development. Neurodevelopmental disorders, or even psychiatric disorders with a neurodevelopmental origin such as schizophrenia27, need better neurobiological characterisation during development28. We focus on two regions with different developmental timelines: the cerebellum and the thalamus. As described above, the cerebellum expands greatly after birth, when the Purkinje cells dendritic trees start growing. This slow maturation exposes the cerebellum longer to developmental insults. In contrast to the cerebellum, the thalamus develops relatively quickly, and its neurons expand and complexify more rapidly2,29 (Figure 1A).

Overview of the rationale of the study.

(A) Context: the cerebellum exhibits protracted growth compared to other brain regions (e.g., thalamus), making it a particularly vulnerable region for neurodevelopmental disorders. Its specific microstructural organisation (i.e. layered with Purkinje cells and granular cells) changes between birth and adulthood. (B) The current “standard” is to probe regional growth with manganese-enhanced MRI (MEMRI). However, it is not sensitive to the underlying microstructural changes. (C) Pushing limits: diffusion-weighted MRS (dMRS) can potentially assess cell-specific microstructural changes in given regions during development. Different measures of metabolites diffusion properties (apparent diffusion coefficient, ADC, at long diffusion times, signal attenuation, S, at high diffusion weighting, b-values) can be interpreted with biophysical modelling and help characterise the nature of the microstructural changes.

By contrasting these two regions, we want to assess the sensitivity of the proposed method. We scanned developing rats from postnatal day five (P5) to P30. We first used MEMRI to confirm the protracted cerebellar development compared to the thalamus. We then acquired diffusion properties of metabolites to probe short- and long-range restriction (i.e., soma restriction and branching). To interpret our data, we applied two biophysical models (one analytical and one computational) to characterise the specific changes in microstructure. We expected monotonic changes in microstructural parameters in both regions, reflecting cell growth, with thalamic properties plateauing earlier than cerebellar properties. This is mostly what we report in our observations. However, we report a U-shape trend for the cerebellar segment length (Lsegment, the distance between two embranchments of a dendritic tree) that matches openly available 3D cell reconstructions. In contrast to common findings in the adult brain, we report that tNAA and Glu do not provide reliable morphological neuronal markers in neonates. We identify instead total creatine (tCr) and taurine (Tau) as more reliable markers for tracking early neuronal development. We believe that this is a promising method to track early neuronal development, particularly in the cerebellum, where microstructure changes drastically postnatally.

Results

Structural imaging confirms protracted cerebellar growth

Representative slices from intra-age non-linear registration are shown in Figure 2A, underlying the cerebellar unfolding. Both cerebellar and thalamic volumes are just below 50µl at P5 (Figure 2B). However, the segmentation at P5 is poor, as explained in Methods (Figure S1), hence data at this time point should be treated cautiously, although the cerebellum and the thalamus, which are medial line structures, are less affected by the atlas distortions.

Manganese-enhanced MRI confirms protracted cerebellar growth in rat pups.

(A) Representative slices around the thalamus and cerebellum of non-linearly averaged images acquired at P5, P15 and P30. (B) Absolute volumetric changes are shown for the segmented thalamus and cerebellum. (C) Normalised volumetric changes. Volume at each age is normalised by the P30 volume.

The cerebellum triples in volume, between P5 and P30, reaching 90% of its P30 volume at P20, while the thalamus reaches 90% of the P30 volume at P15.

Metabolic profile changes with age

The metabolic profile changes noticeably between P5 and P30 as shown in Figure 3A. The absolute concentrations of metabolites, estimated by LCModel using the b0 averaged spectra, shown in Figure 3A, were normalised by the absolute concentration of macromolecules, MM (Figure 3B). Between P5 and P30, metabolite/MM ratios increase for tNAA, Glu, Gln and Ins, and decrease for Tau, Gly, PE and tCho. Note that the increase in tNAA signal with age is protracted in the cerebellum compared to thalamus and has not reached a plateau at P30. These results generally match the metabolic concentrations measured with gold-standard classical MRS in the striatum, cortex and hippocampus of developing rats reported in Reference 30. We observe that tCr/MM increases strongly in the cerebellum between P5 and P30 but appears stable in the thalamus, while Reference 30 reported a 30% increase between P7 and P28 in the 3 ROIs. However, they also reported a 30% increase of the macromolecular content during this developmental period, matching our tCr/MM observation in the thalamus.

Metabolic profile changes postnatally.

(A) Spectra were binned per time point, region and b-value and averaged after phase and frequency correction. The SNR is clearly lower at P5/P10 in both regions: the voxel is about half of the P30 voxel size, and more animals were outliers. The SNR is also lower at P30: the rat size was slightly too big for the mouse cryoprobe at this age, so the skull is not at the top of the probe and the sensitivity is suboptimal (B) Spectra at b=0 ms/µm2 displayed in (A) were quantified with LCModel and absolute metabolite concentrations were normalised by the absolute macromolecular concentration and levelled to get Signal(tCr,P30)/Signal(MM,P30)=8.

Metabolite diffusion properties differ between regions and change with age

Thalamic (Figure 4A) and cerebellar (Figure 4B) signal attenuations up to b=30 ms/µm2 and ADCs at TM=100-1000 ms are shown for tNAA, Glu, tCr, Tau, tCho and Ins. These diffusion properties differ between both regions at each age (see Figure S2 for a region-to-region visual comparison). For most metabolites (except tCho), the signal attenuation is stronger in the thalamus, suggesting less restricted diffusion compared to the cerebellum. There is also a trend for lower signal attenuation and lower ADCs with increasing age, suggesting increasing restriction of metabolite diffusion. This is particularly striking for Tau in the thalamus. To give an estimation of the MRS fit performance, CRLBs are given for tNAA, Glu, tCr, Tau, tCho and Ins for the different diffusion conditions in Table S1.

Biophysical modelling issued from metabolites diffusion properties shows differential developmental trajectories in the cerebellum (A) and in the thalamus (B).

The first column (A,B) reports the signal attenuation as a function of b-value at TM=100ms, and the second column reports the ADC varying with diffusion time. Shadowed error bars represent the standard deviations. Modeling results from the spheres+”astrosticks” model is reported in the third column (A,B). P5 is shaded in the cerebellum due to possible motion artefact. The parameter space (sphere fraction, sphere radius) is displayed. Error bars represent the standard deviations. Llength, modeling output from the morphometric model is shown in the fourth column (A,B). P5 is shaded in the cerebellum due to possible motion artefact.

Biophysical modelling underlines different developmental trajectories of cell microstructure between the cerebellum and the thalamus

Data were fitted with a spheres + “astrosticks” model and a morphometric model (more details in the Methods section). The spheres + “astrosticks” model captures diffusion within spheres (representing cell bodies) and randomly oriented sticks (representing cell processes/neurites): parameters extracted are sphere fraction, fsphere (or equivalently neurite fraction as 1-fsphere) and cell radius, R. The parameter space (sphere radius vs sphere fraction) is represented in the 3rd column of Figure 4A (cerebellum) and Figure 4B (thalamus) for tNAA, Glu, tCr, Tau, tCho and Ins at all ages (see Figure S3 for a region-to-region visual comparison).

Mean values and standard deviations are reported in Table 1, along with the results from the linear mixed effect model ∼ Age + Region + Age:Region + (1|PupID). The fits on averaged data, for binned spectra by age and region, are shown in Figure S4.

The morphometric model, only applied to averaged ADCs at long td, interprets diffusion in branched structures: the segment length, Lsegment is reported on the last column of Figure 4A/B (see Figure S5 for a region-to-region visual comparison and Figure S6 for all parameters estimated by the morphometric model: Dintra, NBranch, Lsegment).

Neuronal markers

tNAA and Glu are primarily concentrated in neurons in adulthood.

Thalamus: The sphere fraction and the radius estimated from tNAA and Glu diffusion properties do not change much with age. For both metabolites, fsphere is around 0.25 and R decreases to 5-6 µm. Lsegment is stable with age for tNAA, but shows a reduction at P15-P20 for Glu.

Cerebellum: The sphere fraction and the radius estimated from tNAA diffusion properties vary with age. fsphere nears 0.5 in the cerebellum and R decreases below 5 µm for tNAA and Glu. Lsegment decreases until P15 for both tNAA and Glu.

Between regions, sphere fraction differs significantly (Table 1).

Glial markers

tCho and Ins are primarily concentrated in glial cells in adulthood. The general trend is a high sphere fraction (∼0.7-0.9) for Ins and moderately high (∼0.5-0.6) for tCho. The radius estimated from tCho differs significantly between regions. Note the trend of decreasing sphere fraction in the cerebellum with age. Data at P15 for Ins in the thalamus were individually not reliable enough to apply the spheres + astrosticks model (c.f. Table S1 and the particularly higher CRLBs at P15 for Ins).

Non-specific markers

tCr and Tau are not known to be compartmentalised in a specific cell-type.

Thalamus: Parameter estimates for tCr diffusion properties do not change substantially with age (fsphere∼0.35, R∼6.5µm). Tau parameters estimation varies a lot with age: R shows a substantial decrease and fsphere increases.

Cerebellum: fsphere decreases from 0.63 (P10) to 0.41 (P30), but R is stable. Tau follows a similar pattern, although with an overall lower fsphere (0.43 at P10 to 0.29 at P30). Lsegment increases in the cerebellum from P10 for tCr and increases P15 for Tau.

Between regions, the sphere fraction differs significantly for both metabolites (Table 1), but only Tau exhibits a significant change with age, and this change is different between regions (significant interaction term). For Tau, the radius is significantly different between regions, changes significantly with age and this change is different between regions (significant interaction term).

Discussion

As in Reference 2 we used MEMRI to validate the protracted development of the cerebellum in rat neonates3, unfolding and tripling volume between P5 and P30. Cell-specific cerebellar and thalamic microstructural features during early development in rat neonates were assessed with dMRS. We specifically report diffusion properties for neuronal markers tNAA and Glu, glial markers tCho and Ins, and the non-specific markers tCr and Tau. Metabolites exhibit significantly different diffusion properties between regions, demonstrating the relevance of dMRS to probe early cell structure development. The sphere fraction is notably higher in the cerebellum for metabolites compartmentalised fully (tNAA, Glu) or partially (tCr, Tau) in neurons, suggesting the cell-body dense granular layer of the cerebellum. The sphere fraction possibly reflects the relative volume of sphere-like structure (e.g., cell bodies) compared to the relative volume of fibrous structures (e.g., cell processes), illustrated in Figure 5A.

The biophysical parameters are interpreted based on the cerebellar microstructural development and main results are summarised.

(A) Schematic of cerebellar neuronal development and biophysical modelling parameters interpretation. Lsegment is interpreted as the distance between two branches of the dendritic tree, NBranch as the number of embranchments. The sphere radius Rsphere is interpreted as the average soma radius (i.e. granule and Purkinje cells). The sphere fraction fsphere represents the volume of sphere-like cell components (e.g., somas) over the volume of fibrous cell components (e.g., cell processes or “neurite”). The sphere fraction is therefore inversely related to the dendritic tree growth. (B) Summary of the cerebellar results (C) Tau seems to change compartments in the thalamus with age, going from neuronal-like compartments (low sphere fraction) to glial-like compartments (high sphere fraction), also indicated by its important signal drop with age. It suggests that Tau is a marker of neuronal maturation. Its concentration remains high up to P30 in the cerebellum. Tau properties in the cerebellum (decreasing sphere fraction + relatively low sphere fractions) and contrast with the thalamus suggest that Tau could be a marker of cerebellar neuronal development.

While soma morphologic maturation mostly happens in the first week31, dendritic expansion is rapid before P15 but then continues to mature up to the end of the 4th week31,32. Remarkably, the sphere fraction extracted from tCr and Tau decreases with age in the cerebellum only, potentially underlining the drastic Purkinje cell growth between birth and adulthood. The diffusion properties of glial metabolites, Ins and tCho, also display a trend to sphere fraction decrease with age. This could be explained by another cerebellar peculiarity, Bergmann glia develop long processes along Purkinje cells and this specific growth could play a role in the sphere fraction decrease of Tau and tCr.

We confirm in this study the observation from Reference 30 that both tCr and Tau have high spectroscopic signals in neonates. They are the two metabolites most reliably quantified up to P15 in both regions (continuing for Tau in the cerebellum up to P30) and the estimates of their diffusion properties are the most confident. The notable decrease of Tau concentration in the thalamus, the extreme sphere radius decrease, and the sphere fraction increase with age extracted from Tau diffusion properties may indicate a change in Tau compartmentation with age in the thalamus. Note that Tau has a higher Dintra33 compared to the other metabolites, potentially affecting the parameter estimation given for Tau. As described in Methods, the fit for taurine was substantially improved for Dintra > 0.6 µm2/ms. The results for Dintra=0.7 µm2/ms are similar for all the other metabolites (Figure S7), but it brings Tau sphere radius estimation to plausible values. However, in all cases, the increase in sphere fraction with age remains: at P5, fsphere(Tau) nears fsphere(tNAA), while at P30 fsphere(Tau) nears fsphere(tCho). We can reasonably hypothesize that Tau is relatively more concentrated in neurons at birth and loses its neuronal compartmentalisation with neuronal maturation, ending up concentrated in glial cells. Moreover, the volume taken by glial cells in the brain is lower than the volume taken by neurons: the decrease in Tau signal matches this proposed switch. Tau is crucial in early development, and it has been shown in cats (who are unable to synthetise Tau, and instead must source it from their diet), that Tau dietary deficiency delays cortical and cerebellar maturation34. Older immunohistochemistry studies associated Tau with Purkinje cells specifically35,36 and Tau levels in the cerebellum are still high at P30, matching the protracted development of Purkinje cells and the cerebellum. Moreover, Tau’s decreasing sphere fraction indicates that it stays relatively highly concentrated in neurons until P30, following the neuronal cerebellar growth and complexification (Figure 5A&C). Given that the low tNAA and Glu signal up to P10 (inclusive), particularly in the cerebellum, prevents confident inference of their diffusion properties, we suggest Tau as a cerebellar marker of early neuronal development, and possibly more specifically to Purkinje cells. Total creatine stands as a good marker too. However, there is no indication of a preferential neuronal compartmentation for this metabolite during early development, hence it is less specific to neurons. This is confirmed by the higher fsphere(tCr) compared to fsphere(Tau) up to P15 in the cerebellum.

We also observed a protracted increase in tNAA signal in the cerebellum compared to the thalamus. tNAA is usually described as a neuronal osmolyte, meaning it could reflect neuronal growth and maturation and be a marker of neuronal development. However, at the voxel level, its signal also reflects the neuronal volumetric contribution, i.e. differences in neuronal densities might mask the effect of development. In contrast, we want to reiterate that dMRS is a normalised measure and dMRS outcomes are not affected by the neuronal density: they represent the average morphological properties of the cell types the metabolite belongs to.

We chose to interpret the dMRS data with two biophysical models: an analytical spheres & “astrosticks” model and a computational morphometric model. The analytical model was applied simultaneously to the signal attenuation at high b-values and the ADCs at long diffusion times, and on individual animals. In addition to the valuable biophysical biomarkers, it provides a good signal representation to evaluate the significance of the differences between diffusion properties of individual datasets. The experimental design was not suitable for a randomly oriented infinite cylinders (“astrocylinders”) model as in previous studies22,23. The mixing time used for the acquisition of signal decay at high b-values (TM=100 ms) is too long to enable sensitivity to fibre radius37, hence the choice of “astrosticks” model. The sphere compartment was added to account for the high sphere fraction in the cerebellum, and for the nature of the neonatal brain (under-developed neurons). The RMSEs from the different models tested (Figure S2) tend to be more similar at P30, suggesting that it is less important to add a sphere compartment for modelling metabolites diffusion in the adult brain.

The computational morphometric model was applied to attempt to extract cell extension-related parameters, but the study design and data quality are suboptimal, and we could only reliably extract Lsegment and Dintra (Figure S6). Dintra is mostly stable with age in both regions. However, it sometimes drops or peaks, likely affecting the other parameters’ estimations. The longest mixing time (TM=1000 ms) is short compared to previous studies using a similar modelling approach21,23, meaning that the diffusion path is shorter (i.e., about 30µm in this study compared to about 45µm in previous studies, assuming Dintra=0.5 µm2/ms). Note that the higher Dintra of Tau is therefore an asset for probing early neuronal cerebellar growth. In our study, the number of embranchments (Nbranch) was very poorly evaluated because of the high level of noise in the data. However, even with a better dataset quality and a more optimal study design, it is unlikely that the Nbranch estimation could reflect the Purkinje cell morphometry because of their size and complexity in adulthood.

Finally, we expected that Lsegment estimated from the morphometric model in the cerebellum would increase regularly with age, matching the dendritic tree development. For neuronal, or partly neuronal metabolites, we report an Lsegment decrease first. Except for tNAA, which consistently decreases in the cerebellum with age, Lsegment decreases until P15 for tCr, Tau and Glu and then increases, forming a U-shape curve. To better understand these results, we extracted the morphometries of developing Purkinje cells and developing thalamic interneurons (both in the mouse) from NeuroMorph.org. Interestingly, from these 3D reconstructed cells, we observe a decrease in Lsegment, 3D-recon up to P13 for Purkinje cells, whilst being quite stable in the thalamus (Figure 6), matching our non-invasive measurement trend. This could reflect the complexification of the dendritic tree following its expansion.

Morphometric parameters extracted from mouse real cells 3D reconstructions on NeuroMorph shows a decrease of Lsegment up to P13 in developing cerebellar Purkinje cells, but not in developing thalamic interneurons.

(A) Examples of cells extracted from NeuroMorph. The total number of cells used per age and region is given in Supplementary Material. References: Cerebellum P6-P1038, P10-P1339, P35-P4340; Thalamus P9-P3041 (B) The values of Branch Length (Lsegment), Branching Order and total process length are directly estimated from the morphology using the function stat_tree in the Trees Matlab Toolbox.

Conclusion

This study proposes diffusion-weighted MRS as a method to longitudinally and non-invasively measure cell-specific cerebellar and thalamic microstructural features during early development in rat neonates. We demonstrate the sensitivity of Taurine to early neuronal cerebellar development. This is a stepping-stone enabling further studies in pathological conditions.

Methods

Animals

Experiments were approved by the United Kingdom Home Office. 18 Sprague-Dawley rats (9 females) were scanned at P5, P10, P15, P20 and P30. Pups came from 3 litters (6 pups scanned per litter). Animals were induced with 4% isoflurane in oxygen and anaesthesia was maintained with 1-2% isoflurane in oxygen. Respiration was monitored with a breathing pillow. Temperature was monitored with a rectal probe, which was used as a cutaneous probe from P5 to P20 to avoid any risk of harm. Isoflurane level was adapted to keep the animals at 50-70 rpm and a warm water-circulation heating pad was used to keep the animals at 36-37°C during scanning. Pups were gently held with foam and protected tape to minimise motion, because the neonatal skull is not suitable for the use of ear bars.

For logistic reasons we could not acquire MEMRI for all litters. Animals from the second litter had manganese-enhanced anatomical scans. To this end, they were intraperitoneally injected with a 50 mg/kg solution of MnCl2 24 hours prior to the scan. Under the age of P10, the dam received the intraperitoneal injection, and the manganese was transferred to the neonates via the maternal milk2.

MRI acquisitions

Data were acquired on a 7T Biospec Bruker MRI scanner (Paravision 360.1, max gradient strength: 660 mT/m), equipped with a mouse quadrature cryoprobe, well-suited for rat neonates. An anatomical scan around the region of interest (thalamus or cerebellum) to place the voxel (T2-weighted, 2D RARE, R=8, 0.2 mm isotropic resolution, TE=33 ms, TR=3500 ms). For 6 animals, this scan was replaced by a whole brain manganese-enhanced anatomical scan (T1-weighted, 3D MGE, TE=2.4 s, TR=62 ms, 0.15 mm isotropic resolution).

dMRS acquisitions

A STELASER (Ligneul et al., MRM 2017) sequence (TE=34 ms, diffusion gradient duration, δ=5 ms) was used for dMRS acquisitions. The voxel was placed in the cerebellum for 9 pups, and in the thalamus for the other 9 pups. Voxel size was increased with age to follow brain growth. Diffusion-weighted spectra were acquired at 4 mixing times (TM) and 2 b-values (c.f. Table 2 for detailed parameters): a small “b0” used as a crusher for the stimulated echo, and b0+3 ms/µm2. Higher b-values were also acquired at the shortest diffusion time to probe restriction within cell bodies and enable inference of cell fraction and radius.

Processing

dMRS

Processing was achieved with a custom Matlab (The Mathworks) routine (https://github.com/clemoune/dmrs-neonates/tree/main/A_dMRS_processing):

Signals from different coils were combined, using the water signal at the lowest b-value and shortest diffusion time.

Outliers were identified and discarded, as described in42.

For each transient i, the mean absolute upfield signal (range 1.9-4.5 ppm), Mup(i) was computed.

Transients affected by motion are characterised by a clear drop of Mup(i). Transients falling below the median(Mup) - 3xMAD threshold were discarded.

Experiments at high b-values are more susceptible to changing gradient direction (particularly if the voxel contains white matter tracts). To predict its effect across b-values and overcome the small number of acquisitions per unique condition (i.e. a (b-value, direction) pair), a diffusion kurtosis tensor was computed based on Mup. The MAD calculated at b=0 was used as a “ground truth” MAD and was reported to higher b-values (per direction). High b-values acquisitions are more easily corrupted by motion and therefore present an artificially high MAD.

Individual transients were corrected for phase and frequency drifts and averaged across all diffusion directions (powder-averaged) via an arithmetic mean. In general, metabolites signal was visually detectable on single transients (with a line-broadening of 5 Hz), except at b=20,30 ms/µm2 at P5, rendering these acquisitions more susceptible to motion artefacts.

Any water residual was removed using a singular value decomposition method centred around the water region. If the baseline was heavily distorted despite this step due to very poor water suppression, the acquisition was discarded from the analysis.

Spectra were subsequently fitted with LCModel. The simulated basis set contained the following metabolites: Ace, Ala, Asp, Cr, GABA, Gln, Glu, Gly, GPC, GSH, Ins, sIns, Lac, NAA, NAAG, PCho, PCr, Tau, PE, and included a macromolecules (MM) baseline acquired experimentally (double-inversion recovery and b=10 ms/µm2).

Diffusion properties of NAA+NAAG (tNAA), Glu, Cr+PCr (tCr), Tau, GPC+PCho (tCho) and Ins are reported. In order to model both signal compartmentalisation and restricted diffusion by cell membranes, we investigated metabolites diffusion through two sets of measurements: i) the direction-averaged signal decay as a function of the diffusion weighting b, keeping the diffusion time constant at 104.25 ms (time between diffusion gradients Δ fixed at 105.91 ms and δ fixed at 5 ms) and varying the strength of the diffusion-sensitizing gradients; and ii) the apparent diffusion coefficient (ADC) as a function of increasing diffusion time (keeping δ fixed at 5 ms). For the measurements i), data are reported as the ratio S(b)/S(b=0). For the measurements ii), ADCs diffusion time dependencies are reported. ADCs are calculated following Negative ADCs were systematically removed as they point to a quantification error.

In the cerebellum, metabolites’ signal attenuation at P5 is always much stronger and quite far off the other time points. The macromolecules’ signal attenuation at high b-values is supposed to remain low43 and can be used as an internal probe for motion artefact33. However, the macromolecules’ signal attenuation is quite strong at P5 in the cerebellum (Figure S8), pointing out to a residual motion artefact despite careful processing. The spectroscopic signal in individual transients was too low at high b-values, notably in the cerebellum: it is possible that the spectral realignment failed.

MEMRI

The three first echoes of each MGE acquisition were averaged and denoised using the DenoiseImage function from the MINC toolbox, based on44. For each time point, an image with no obvious artefact and a good visual CNR was taken as the initial model and transformed into the Fischer atlas45 space. Images were then processed using registration_tamarack.py with the MAGeT segmentation in the available Pydpider pipelines46. Due to the large deformations between P30 and P5, the P30 timepoint could not be used as a reference. An adequate atlas from P15 was generated in a couple of iterations (using a P20 atlas as intermediary). It was then used as a reference atlas and reference timepoint for the pipeline. The olfactory bulb and the more frontal parts were poorly registered because the image often had an intensity drop there, but it did not affect the registration of caudal regions, such as the cerebellum and the thalamus. Image with strong intensity differences, or corrupted by motion were removed from the analysis. The registration did not work well at P5, even when attempting to use an atlas generated at P10 (that is already slightly distorted).

Analysis

Signal decays at high-b values (at TM=100 ms) and ADCs at varying diffusion times were fitted together with a spheres and “astrosticks” (corresponding to randomly oriented sticks) model using the functions implemented in DIVE (https://git.fmrib.ox.ac.uk/fsl/DIVE). We fitted the signal to: Free parameters were R (sphere radius) and fsphere (sphere fraction). Dintra (i.e. the metabolite intracellular free diffusivity) was fixed for both compartments at 0.5 µm2/ms. Previous studies measuring brain metabolites diffusion properties at ultra-short diffusion-times33,47 estimated the intracellular free diffusivity at 0.5-0.6 µm2/ms for most metabolites (except Tau). However other studies using different (less direct) approaches reported lower Dintra21,22, hence we tested a range of reasonable Dintra. Dintra was fixed 0.3, 0.4, 0.5, 0.6, 0.7 and 0.8 µm2/ms. RMSEs between model fit and data were equivalent for all Dintra, except for Tau in the thalamus where increasing Dintra improved the fit, particularly until Dintra=0.5µm2/ms. Moreover, parameters estimation was unrealistic for low Dintra. Increasing Dintra between 0.5 µm2/ms and 0.7 µm2/ms did not change the trend of the results, except for Tau in the thalamus (see Figure S0 and Discussion). The fit was made robust to missing data (when excluded during processing).

To test for the model accounting best for our data, we tested different analytical models. RMSE between model fit and data were calculated for “astrosticks” only (1 free parameter, Dintra), “astrocylinders” (corresponding to randomly oriented infinite cylinders, 2 free parameters: Dintra and cylinder radius), spheres and astrosticks without constraint (3 free parameters) and spheres and astrosticks with Dintra fixed at 0.5 µm2/ms (Figure S0). RMSE were the lowest for spheres and astrosticks, and were very similar with and without the Dintra constraint. Hence the model with the least free parameters was chosen. Moreover, the results from the constrained spheres and astrosticks model were more stable and physically meaningful with the constraint on Dintra. The code used for the analysis can be found on: https://github.com/clemoune/dmrs-neonates/tree/main/B_Astrosticks_Spheres_model.

Long diffusion times

Average ADCs at varying diffusion times were fitted with the morphometric model introduced in21. Modeling was performed using the dictionary generated for23 with TM = 100, 750 ms predicted by interpolation, and a random forest regressor (‘treeBagger’ function in MatlabR2020b, with tree depth = 20; number of trees = 200 and other model hyper-parameters as per default) trained in a fully supervised fashion, using the ADCs time dependencies as input and the ground truth morphological parameters as targets; as loss function the mean squared error between predicted and ground truth parameters values was used. Parameters fitted were Dintra, Nbranch, Lsegment, SDNbranch and SDLsegment. Boundaries for fitting the model to data were respectively [0.1, 1], [2, 25], [5, 100], [2, 3] and [5, 10]. Given the data noise, only Lsegment and Dintra are estimated robustly. Fits for all parameters are shown in Supplementary Information.

Statistical analyses

Statistical analyses were run with R, using the “lmerTest48 package. To assess whether diffusion properties were significantly different in the 2 regions during development, we applied a linear mixed effect model to the parameters extracted from the spheres + “astrosticks” model (sphere fraction and sphere radius), using age and region as fixed effects, and a random intercept for pups ID. An analysis of variance enabled to identify whether the inclusion of an interaction term age*region was necessary or not. Results were corrected for multiple hypotheses testing with a Bonferroni correction (6 metabolites,2 parameters = 12 hypotheses). The threshold corrected p-value is 4.2e-3 for α=0.05 and 8.3e-4 for α=0.01.

Supporting information

eLife Assessment

This study presents valuable findings regarding the microstructural basis of brain development in the cerebellum and thalamus of rat neonates using diffusion-weighted MRS. The authors present solid evidence of differential development trajectories in the thalamus and the cerebellum through analytical and morphometric biophysical modelling of the diffusion-weighted MRS data, though some aspects such as the validation of the findings against gold-standard techniques and a detailed discussion of methodological choices require further elaboration. The work will be of interest to developmental biologists and neuroscientists seeking noninvasive approaches to probe in vivo neuronal and glial development in the brain.

Reviewer #1 (Public Review):

In this work, Ligneul and coauthors implemented diffusion-weighted MRS in young rats to follow longitudinally and in vivo the microstructural changes occurring during brain development. Diffusion-weighted MRS is here instrumental in assessing microstructure in a cell-specific manner, as opposed to the claimed gold-standard (manganese-enhanced MRI) that can only probe changes in brain volume. Differential microstructure and complexification of the cerebellum and the thalamus during rat brain development were observed non-invasively. In particular, lower metabolite ADC with increasing age were measured in both brain regions, reflecting increasing cellular restriction with brain maturation. Higher sphere (representing cell bodies) fraction for neuronal metabolites (total NAA, glutamate) and total creatine and taurine in the cerebellum compared to the thalamus were estimated, reflecting the unique structure of the cerebellar granular layer with a high density of cell bodies. Decreasing sphere fraction with age was observed in the cerebellum, reflecting the development of the dendritic tree of Purkinje cells and Bergmann glia. From morphometric analyses, the authors could probe non-monotonic branching evolution in the cerebellum, matching 3D representations of Purkinje cells expansion and complexification with age. Finally, the authors highlighted taurine as a potential new marker of cerebellar development.

From a technical standpoint, this work clearly demonstrates the potential of diffusion-weighted MRS at probing microstructure changes of the developing brain non-invasively, paving the way for its application in pathological cases. Ligneul and coauthors also show that diffusion-weighted MRS acquisitions in neonates are feasible, despite the known technical challenges of such measurements, even in adult rats. They also provide all necessary resources to reproduce and build upon their work, which is highly valuable for the community.

From a biological standpoint, claims are well supported by the microstructure parameters derived from advanced biophysical modelling of the diffusion MRS data. The assumption of metabolite compartmentation, forming the basis of cell-specific microstructure interpretation of dMRS data, remains debated and should be considered with care (Rae, Neurochem Res, 2014, https://doi.org/10.1007/s11064-013-1199-5). External cross-validation of some of the authors' claims, in particular taurine in the thalamus switching from neurons to astrocytes during brain development, would be a highly valuable addition to this study.

Specific strengths:

(1) The interpretation of dMRS data in terms of cell-specific microstructure through advanced biophysical modelling (e.g. the sphere fraction, modelling the fraction of cell bodies versus neuronal or astrocytic processes) is a strong asset of the study, going beyond the more commonly used signal representation metrics such as the apparent diffusion coefficient, which lacks specificity to biological phenomena.

(2) The fairly good data quality despite the complexity of the experimental framework should be praised: diffusion-weighted MRS was acquired in two brain regions (although not in the same animals) and longitudinally, in neonates, including data at high b-values and multiple diffusion times, which altogether constitutes a large-scale dataset of high value for the diffusion-weighted MRS community.

(3) The authors have shared publicly data and codes used for processing and fitting, which will allow one to reproduce or extend the scope of this work to disease populations, and which goes in line with the current effort of the MR(S) community for data sharing.

Specific weaknesses:

(1) This work lacks an introduction and a discussion about diffusion MRI, which is already a validated technique to assess brain development non-invasively. Although water lacks cell-specificity compared to metabolites, several studies have reported a decrease in water ADC and increased fractional anisotropy with brain maturation, associated with the myelination process and decreased water content (overview in Hüppi, Chapt. 30 of "Diffusion MRI: Theory, Methods, and Applications", Oxford University Press, 2010). Interestingly, the same observations are found in this work (decreased ADC with age for most metabolites in both brain regions), which should have been commented on. Moreover, the authors could have reported water diffusion properties in addition to metabolites', as I believe the water signal, used for coil combination and/or Eddy currents corrections, is usually naturally acquired during diffusion-weighted MRS scans.

(2) It is unclear why the authors have normalized metabolite concentrations (measured from low b-values diffusion-weighted MRS spectra) to the macromolecule concentrations. First, it is not specified whether in vivo macromolecules were acquired at each age or just at one time point. Second, such ratios are not standard practice in the MRS community so this choice should have been explained. Third, the macromolecule content was reported to change with age (Tkac et al., Magn Reson Med, 2003), therefore a change in metabolite to macromolecule ratio with age cannot be interpreted unequivocally.

(3) Some discussion is missing about the choice of the analytical biophysical model (although a few are compared in Supplementary Materials), in particular: is a model of macroscopic anisotropy relevant in cerebellum, made of a large fraction of oriented white matter tracks, and does the model remain valid at different ages given white matter maturation and the ongoing myelination process?

Reviewer #2 (Public Review):

Summary:

The authors set out to non-invasively track neuronal development in rat neonates, which they achieved with notable success. However, the direct relationship between the results and broader conclusions regarding developmental biology and potential human implications is somewhat overstretched without further validation.

Strengths:

If adequately revised and validated, this work could have a significant impact on the field, providing a non-invasive tool for longitudinal studies of brain development and neurodevelopmental disorders in preclinical settings.

Weaknesses:

(1) Consistency and Logical Flow:

- The manuscript suffers from a lack of strategic flow in some sections. Specifically, transitions between major findings and methodological discussions need refinement to ensure a logical progression of ideas. For example, the jump from the introduction of developmental trajectories and the technicalities of MRS (Magnetic Resonance Spectroscopy) processing on page 3 could benefit from a bridging paragraph that explicitly states the study's hypotheses based on existing literature gaps.

(2) Scientific Rigour:

- While the novel application of diffusion-weighted MRS is commendable, there's a notable gap in the rigorous validation of this approach against gold-standard histological or molecular techniques. Particularly, the assertions regarding the sphere fraction and morphological changes inferred from biophysical modelling mandates direct validation to solidify the claims made. A study comparing these in vivo findings with ex vivo confirmation in at least a subset of samples would significantly enhance the reliability of these conclusions.

(3) Clarity and Novelty:

- The manuscript often delves deeply into technical specifics at the expense of accessibility to readers not deeply familiar with MRS technology. The introduction and discussions would benefit from a clearer elucidation of why these specific metabolite markers were chosen and their known relevance to neuronal and glial cells, placing this in the context of what is novel compared to existing literature.

- The novelty aspect could be reinforced by a more structured discussion on how this method could change the current understanding or practices within neurodevelopmental research, compared to the current state of the art.

(4) Completeness:

- The Discussion section requires expansion to offer a more comprehensive interpretation of how these findings impact the broader field of neurodevelopment and psychiatric disorders. Specifically, the implications for human studies or clinical translation are touched upon but not fully explored.

- Further, while supplementary material provides necessary detail on methodology, key findings from these analyses should be summarized and discussed in the main text to ensure the manuscript stands complete on its own.

(5) Grammar, Style, Orthography:

- There are sporadic grammatical and typographical errors throughout the text which, while minor, detract from the overall readability. For example, inconsistencies in metabolite abbreviations (e.g., tCr vs Cr+PCr) should be standardized.

(6) References and Additional Context:

- The current reference list is extensive but lacks integration into the narrative. Direct comparisons with existing studies, especially those with conflicting or supportive findings, are scant. More dedicated effort to contextualize this work within the existing body of knowledge would be beneficial.

Author response:

The following is the authors’ response to the original reviews

Summary of revisions:

Thanks to the careful review and comments from the reviewers, we restructured the introduction and the discussion to improve clarity and better contextualise findings. We notably discuss further the fsphere decrease observations in the cerebellum and the Tau-specific findings (Tau being a possible marker for Purkinje cells development and Tau switching compartment in the thalamus). We added material in Supplementary Information to support these discussion points. We added a figure to show the metabolic profiles normalised by water or by macromolecules and a figure and table related to a rough approximation of fsphere, leaning on existing literature. We report the DTI results for thoroughness.

Public Reviews:

Reviewer #1 (Public Review):

In this work, Ligneul and coauthors implemented diffusion-weighted MRS in young rats to follow longitudinally and in vivo the microstructural changes occurring during brain development. Diffusion-weighted MRS is here instrumental in assessing microstructure in a cell-specific manner, as opposed to the claimed gold-standard (manganese-enhanced MRI) that can only probe changes in brain volume. Differential microstructure and complexification of the cerebellum and the thalamus during rat brain development were observed noninvasively. In particular, lower metabolite ADC with increasing age were measured in both brain regions, reflecting increasing cellular restriction with brain maturation. Higher sphere (representing cell bodies) fraction for neuronal metabolites (total NAA, glutamate) and total creatine and taurine in the cerebellum compared to the thalamus were estimated, reflecting the unique structure of the cerebellar granular layer with a high density of cell bodies. Decreasing sphere fraction with age was observed in the cerebellum, reflecting the development of the dendritic tree of Purkinje cells and Bergmann glia. From morphometric analyses, the authors could probe non-monotonic branching evolution in the cerebellum, matching 3D representations of Purkinje cells expansion and complexification with age. Finally, the authors highlighted taurine as a potential new marker of cerebellar development.

From a technical standpoint, this work clearly demonstrates the potential of diffusion-weighted MRS at probing microstructure changes of the developing brain non-invasively, paving the way for its application in pathological cases. Ligneul and coauthors also show that diffusionweighted MRS acquisitions in neonates are feasible, despite the known technical challenges of such measurements, even in adult rats. They also provide all necessary resources to reproduce and build upon their work, which is highly valuable for the community.

From a biological standpoint, claims are well supported by the microstructure parameters derived from advanced biophysical modelling of the diffusion MRS data. The assumption of metabolite compartmentation, forming the basis of cell-specific microstructure interpretation of dMRS data, remains debated and should be considered with care (Rae, Neurochem Res, 2014, https://doi.org/10.1007/s11064-013-1199-5). External cross-validation of some of the authors' claims, in particular taurine in the thalamus switching from neurons to astrocytes during brain development, would be a highly valuable addition to this study.

R1.1: We understand the reviewer's concerns. Metabolic compartmentation is not a one-toone correspondence. Although we interpret the results in the light of metabolic compartmentation, our results are not driven by this assumption. We could not perform a direct cross-validation of the taurine switch in the thalamus, but we now clarify in the discussion why the dMRS results themselves indicate a switch, and we integrate our results better with existing literature on taurine. We now discuss this in more detail for the cerebellar results too.

Specific strengths:

(1) The interpretation of dMRS data in terms of cell-specific microstructure through advanced biophysical modelling (e.g. the sphere fraction, modelling the fraction of cell bodies versus neuronal or astrocytic processes) is a strong asset of the study, going beyond the more commonly used signal representation metrics such as the apparent diffusion coefficient, which lacks specificity to biological phenomena.

(2) The fairly good data quality despite the complexity of the experimental framework should be praised: diffusion-weighted MRS was acquired in two brain regions (although not in the same animals) and longitudinally, in neonates, including data at high b-values and multiple diffusion times, which altogether constitutes a large-scale dataset of high value for the diffusion-weighted MRS community.

(3) The authors have shared publicly data and codes used for processing and fitting, which will allow one to reproduce or extend the scope of this work to disease populations, and which goes in line with the current effort of the MR(S) community for data sharing.

Specific weaknesses:

(1) This work lacks an introduction and a discussion about diffusion MRI, which is already a validated technique to assess brain development non-invasively. Although water lacks cellspecificity compared to metabolites, several studies have reported a decrease in water ADC and increased fractional anisotropy with brain maturation, associated with the myelination process and decreased water content (overview in Hüppi, Chapt. 30 of "Diffusion MRI: Theory, Methods, and Applications", Oxford University Press, 2010). Interestingly, the same observations are found in this work (decreased ADC with age for most metabolites in both brain regions), which should have been commented on. Moreover, the authors could have reported water diffusion properties in addition to metabolites', as I believe the water signal, used for coil combination and/or Eddy currents corrections, is usually naturally acquired during diffusion-weighted MRS scans.

R1.2: Thank you for these helpful suggestions. We have now improved our introduction of the various modalities, and we contextualise the study in light of previous DTI findings in the as suggested by the reviewer. We agree with the reviewer that the comparison with previous human DTI is relevant, and we now mention it at the beginning of the discussion. However, the very different nature of the dMRS signal compared to dMRI (intracellular and absence of exchange for metabolites) prevents us from drawing any strong conclusions.

(2) It is unclear why the authors have normalized metabolite concentrations (measured from low b-values diffusion-weighted MRS spectra) to the macromolecule concentrations. First, it is not specified whether in vivo macromolecules were acquired at each age or just at one time point. Second, such ratios are not standard practice in the MRS community so this choice should have been explained. Third, the macromolecule content was reported to change with age (Tkac et al., Magn Reson Med, 2003), therefore a change in metabolite to macromolecule ratio with age cannot be interpreted unequivocally.

R1.3: We agree with the reviewer that this needed further explanations. We now clarify in the Results section “Metabolic profile changes with age” the reasoning behind choosing macromolecules for normalisation. We also added in the Supplementary Information the metabolite concentrations change with age when normalising by water, and a direct comparison with MM normalisation (Figure S2).

(3) Some discussion is missing about the choice of the analytical biophysical model (although a few are compared in Supplementary Materials), in particular: is a model of macroscopic anisotropy relevant in cerebellum, made of a large fraction of oriented white matter tracks, and does the model remain valid at different ages given white matter maturation and the ongoing myelination process?

R1.4: We agree with the reviewer that this is a valid concern. We actually acquired some standard DTI at the end of the acquisition sessions (where possible) having in mind the fibre dispersion estimation. However, data could not be acquired in all animals, and the data quality was poor (see Figure S8, the experimental conditions would have required further optimisation). We now add a couple of sentences at the beginning and in the end of discussion to address this limitation, and we include the DTI data in Supplementary Information.

Reviewer #2 (Public Review):

Summary:

The authors set out to non-invasively track neuronal development in rat neonates, which they achieved with notable success. However, the direct relationship between the results and broader conclusions regarding developmental biology and potential human implications is somewhat overstretched without further validation.

Strengths:

If adequately revised and validated, this work could have a significant impact on the field, providing a non-invasive tool for longitudinal studies of brain development and neurodevelopmental disorders in preclinical settings.

Weaknesses:

(1) Consistency and Logical Flow:

The manuscript suffers from a lack of strategic flow in some sections. Specifically, transitions between major findings and methodological discussions need refinement to ensure a logical progression of ideas. For example, the jump from the introduction of developmental trajectories and the technicalities of MRS (Magnetic Resonance Spectroscopy) processing on page 3 could benefit from a bridging paragraph that explicitly states the study's hypotheses based on existing literature gaps.

R2.1: Thank you for this general feedback (along with your point (3)) that helped us restructure the introduction and the discussion to improve the clarity and flow.

(2) Scientific Rigour:

While the novel application of diffusion-weighted MRS is commendable, there's a notable gap in the rigorous validation of this approach against gold-standard histological or molecular techniques. Particularly, the assertions regarding the sphere fraction and morphological changes inferred from biophysical modelling mandates direct validation to solidify the claims made. A study comparing these in vivo findings with ex vivo confirmation in at least a subset of samples would significantly enhance the reliability of these conclusions.

R2.2: We agree with the reviewer that this would have been a great addition to the manuscript. Although we could not run new experiments to address these flaws, we now discuss the results more quantitatively, leaning on existing literature (addition of Figure S11 and Table S2). This helps us understand the results around Tau in both regions better, and illustrate the Rsphere trend.

(3) Clarity and Novelty:

- The manuscript often delves deeply into technical specifics at the expense of accessibility to readers not deeply familiar with MRS technology. The introduction and discussions would benefit from a clearer elucidation of why these specific metabolite markers were chosen and their known relevance to neuronal and glial cells, placing this in the context of what is novel compared to existing literature.

- The novelty aspect could be reinforced by a more structured discussion on how this method could change the current understanding or practices within neurodevelopmental research, compared to the current state of the art.

R2.3: See answer to (1). By restructuring the introduction and the discussion, we hope to have addressed this point. We now discuss how these findings compare to the state of the art (notably added comparison with dMRI research). Along with the next comment, we better discuss potential implications of these findings for neurodevelopmental research.

(4) Completeness:

- The Discussion section requires expansion to offer a more comprehensive interpretation of how these findings impact the broader field of neurodevelopment and psychiatric disorders. Specifically, the implications for human studies or clinical translation are touched upon but not fully explored.

Further, while supplementary material provides necessary detail on methodology, key findings from these analyses should be summarized and discussed in the main text to ensure the manuscript stands complete on its own.

R2.4: Thank you for these helpful suggestions. We now integrate the findings better into the existing literature. We notably discuss how the results might translate to humans.

(5) Grammar, Style, Orthography:

There are sporadic grammatical and typographical errors throughout the text which, while minor, detract from the overall readability. For example, inconsistencies in metabolite abbreviations (e.g., tCr vs Cr+PCr) should be standardized.

R2.5: Thank you for the careful review. This has been corrected.

(6) References and Additional Context:

The current reference list is extensive but lacks integration into the narrative. Direct comparisons with existing studies, especially those with conflicting or supportive findings, are scant. More dedicated effort to contextualize this work within the existing body of knowledge would be beneficial.

R2.6: Because the nature of this work is novel, it is difficult to find directly conflicting/similar works. However, we now integrate the findings into the broader literature.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Minor comments:

Thank you for the careful review, we have addressed most of the minor comments, except for the last one, which we discuss below.

- Some figures appear blurred in the printed PDF- Introduction: "constrained and hindered by cell membranes," - maybe use "restricted" instead of "constrained", like everywhere else in the text

- Introduction: "(typically ~8cm3 vs ~8mm3 in dMRI in humans)" - here I suggest to put the rat brain sizes instead to help the reader understand how small the voxel was at P5 in this study, thus explaining the challenges

- Fig 1 - numbers 1 and 2 on panel A,B should be clarified and they do not match 1 and 2 on panel C, which is confusing- Fig 2 - I am guessing the large dots are the mean and small are individual data points? Please clarify

- Please specify "Relative CRLB" rather than just "CRLB", in supp. mat as well

- Fig 3 - title of panel B, I would change "signal" into "concentration"

- Fig 3 - end of caption: "and levelled to get Signal(tCr,P30)/Signal(MM,P30)=8", I think "in the thalamus" is missing

- The results section "Biophysical modelling underlines different developmental trajectories of cell microstructure between the cerebellum and the thalamus" is sometimes unprecise, e.g.: "Cerebellum: The sphere fraction and the radius estimated from tNAA diffusion properties vary with age." but the tNAA sphere fraction seems to vary more with age in the thalamus according to table 1 "Cerebellum: fsphere decreases from 0.63 (P10) to 0.41 (P30), but R is stable" this is for tCr I presume

- Table 1 - "pvalues" please add "before multiple comparison correction"

- Figure 5 - Panel B, the L-segment subpanel is unclear -which metabolites is it referring to? Why does Tau have a * in panel A?

- Update Ref 37 to the journal version

- Methods: "A STELASER (Ligneul et al., MRM 2017) sequence", add numbered reference instead

- Please specify that the DIVE toolbox uses Gaussian phase distribution approximation, it is important for the dMRS reader given that your diffusion gradient length is long and cannot be neglected, and that the SGP approximation does not apply.

The Gaussian phase distribution approximation and the SGP approximation are two different concepts. The gradient duration ∂ (7 ms) is short compared to the gradient separation ∆ (100 ms), but it could still be considered too long for the SGP approximation to hold. However, the gradient duration is accounted for in DIVE in any case.

Version 2

Abstract

Abstract

It is currently impossible to non-invasively assess cerebellar cell structure during early development. Here we propose a novel approach to non-invasively and longitudinally track cell-specific development using diffusion-weighted magnetic resonance spectroscopy in combination with microstructural modelling. Tracking metabolite diffusion allows us to probe cell-specific developmental trajectories in the cerebellum and thalamus of healthy rat neonates from post-natal day (P) 5 to P30. Additionally, by comparing different analytical and biophysical microstructural models we can follow the differential contribution of cell bodies and neurites during development. The thalamus serves as a control region to assess the sensitivity of our method to microstructural differences between the regions. We found significant differences between cerebellar and thalamic metabolites diffusion properties. For most metabolites, the signal attenuation is stronger in the thalamus, suggesting less restricted diffusion compared to the cerebellum. There is also a trend for lower signal attenuation and lower ADCs with increasing age, suggesting increasing restriction of metabolite diffusion. This is particularly striking for taurine in the thalamus. We use biophysical modelling to interpret these differences. We report a decreased sphere fraction (or an increased neurite fraction) with age for taurine and total creatine in the cerebellum, marking dendritic growth. Surprisingly, we also report a U-shape trend for segment length (the distance between two embranchments in a dendritic tree) in the cerebellum agreeing with age-matching morphometry of openly available 3D-Purkinje reconstructions. Results demonstrate that diffusion-weighted MRS probes early cerebellar neuronal development non-invasively.

Introduction

The cerebellum is a well-conserved structure across species and particularly across mammals. The cerebellar cortex is a highly folded structure organized in three layers: the molecular layer (containing the notable Purkinje cell dendritic trees), the Purkinje layer (containing the somas of Purkinje cells), and the granular layer (containing the granule cells). Projections to the deep cerebellar nuclei form the cerebellar white matter. Despite its small size (compared to the cerebrum), it contains more than half of the total brain neurons: this is mainly due to the granule cells, very small and densely packed cells, some of the smallest of the brain, presenting a very simple morphology1. On the other hand, the Purkinje cells are one of the largest types of neurons, and most complex cells in the brain. Cerebellar development is protracted2, starting from birth (or the last trimester of pregnancy) to a few weeks old in rodents (or 2 y.o. in humans)3. The cerebellum is involved in many basic and high-level functions, from motor control to social skills, and it is known to be affected in a number of neurodevelopmental disorders4.

During early development, neurons arborise. Their morphology becomes more complex and exhibits denser branching. In rodents, Purkinje cells’ dendritic trees start growing at birth. In 30 days, they evolve from a simple soma (~10 µm diameter) to a large cell with a highly branched and planar dendritic tree (soma diameter ~20 µm, area ~10,000 µm2 in the mouse5). Purkinje cell morphology can be altered in autism spectrum disorders (ASD), as shown post-mortem, in humans6 and multiple mouse models of ASD. Specifically, they show an increased spine density in Tsc1 mutant mice7, an overgrowth of the dendritic tree and abnormal branching in PTEN KO mice8, a reduction of the dendritic tree in AUTS2 mice9; and general alterations in developmental disorders10–13.

All cerebellar outputs transit via the deep cerebellar nuclei (DCN), and a couple of them project to the thalamus before spreading to the cortex. The thalamus is considered a relay station in the brain and contains about 60 nuclei on unique pathways. Some thalamic nuclei are involved in psychiatric disorders, and more specifically, the microstructure of the mediodorsal nucleus and the pulvinar nucleus are altered in schizophrenia and psychosis14. Some other nuclei are involved in neurodevelopmental disorders (e.g., the lateral reticular nuclei in ASD15). Generally, these disorders are associated with an altered thalamic connectivity14. Thalamocortical projections are crucial for the cytoarchitectural development of the cortex16,17.

Neurodevelopmental disorders, or even psychiatric disorders with a neurodevelopmental origin such as schizophrenia18, generally need better neurobiological characterisation during development19.

Rodents reach adulthood after a couple of months and are an ideal model for mammal brain development. While brain cell developmental trajectories are key to assessing neurodevelopment, there is a lack of translational techniques allowing for a non-invasive measure of their properties. Typically, EEG and MEG signals contain brain activity cell-specific information. Still, it is hard to retrieve in practice, and the existing forward or inverse models notably use prior information on cell morphology. PET can be cell-specific and provide information about metabolism, but it is blind to cell morphology and relies on radioactive ligands, making it very expensive and potentially harmful. fNIRS and ultrasounds are too indirect to measure cell-specific information as they rely on the hemodynamic response. Last but not least, MRI is non-invasive and non-harmful. The MR signal can be sensitised to assess different tissue properties. In the following paragraphs, we will summarise the importance and limitations of some MR methods sensitive to brain structure and microstructure during development and introduce our primary method, namely diffusion-weighted MR spectroscopy (dMRS).

(Structural MRI: anatomy) Manganese-enhanced MRI (MEMRI) is particularly suited to studying whole-brain early development in rodents. It provides high structural contrast and a fast acquisition time in neonates2,20. Still, the method is blind to the microstructural changes that lead to the volumetric changes reported.

(dMRI: microstructure) Diffusion-weighted MRI (dMRI) measures the diffusion properties of water in the brain: restricted and hindered by cell membranes, water diffusion provides an indirect measure of the tissue microstructure21. dMRI is versatile and applied in numerous applications, including the characterisation of the microstructural organisation during neurodevelopment in rodents and, to a lesser extent, in humans. Comparison with postmortem microscopy shows a correlation between a high cortical fractional anisotropy (FA) at birth and the presence of radial glia. The latter progressively disappears in the first week along with an FA decrease, illustrating the potential of dMRI to characterise neurodevelopment22,23. In the cerebellum, the dense folding limits diffusion imaging to exploring white matter or ex vivo samples24.

(dMRS: cell-specific microstructure) Water is ubiquitous and can only partially disentangle the contribution from different tissue compartments (e.g. intra and extracellular space). Diffusion-weighted magnetic resonance spectroscopy (dMRS) measures the diffusion properties of metabolites. Brain metabolites are mostly intracellular, with some more specific to neurons (e.g., N-acetyl-aspartate, NAA, and glutamate, Glu) and others more specific to glial cells (e.g., choline compounds, tCho, and myo-inositol, Ins), therefore different cellular compartments can be resolved25. However, metabolites are 104-fold less concentrated than water, requiring large measurement volumes for dMRS (typically ~ 50-100mm3 in rodents), and long acquisition times. As for dMRI, some dMRS acquisition parameters, such as the diffusion weighting and diffusion time, enable the signal to be sensitised to different cellular morphological scales (e.g., cellular processes, cell bodies, cell extension). Using multiple diffusion times and a post-hoc model, cellular morphological parameters can be estimated from dMRS26. They correlate with neuronal and glial morphologies extracted from histological measures in the healthy, adult mouse27,28. More specifically, it has been shown that myo-inositol diffusion properties reflect astrocytic morphology in an induced model of astrocytic hypertrophy29, and in a mouse model of demyelination where astrocytic activation was expected30. Choline compound’s diffusion properties were associated with microglial activation in human disease31, or in an LPS-induced brain inflammation model32. However, to our knowledge, very few studies have used dMRS to probe changes in neuronal morphology, and none have attempted to probe neuronal development in early life.

Here, we propose a non-invasive method to monitor cellular brain development. We focus on two regions with different developmental timelines: the cerebellum and the thalamus. As described above, the cerebellum expands greatly after birth, when the Purkinje cells dendritic trees start growing. This slow maturation exposes the cerebellum longer to developmental insults. In contrast to the cerebellum, the thalamus develops relatively quickly, and its neurons expand and complexify more rapidly2,33 (Figure 1A).

Overview of the rationale of the study.

(A) Context: the cerebellum exhibits protracted growth compared to other brain regions (e.g., thalamus), making it a particularly vulnerable region for neurodevelopmental disorders. Its specific microstructural organisation (i.e. layered with Purkinje cells and granular cells) changes between birth and adulthood. (B) The current “standard” is to probe regional growth with manganese-enhanced MRI (MEMRI). However, it is not sensitive to the underlying microstructural changes. (C) Pushing limits: diffusion-weighted MRS (dMRS) can potentially assess cell-specific microstructural changes in given regions during development. Different measures of metabolites diffusion properties (apparent diffusion coefficient, ADC, at long diffusion times, signal attenuation, S, at high diffusion weighting, b-values) can be interpreted with biophysical modelling and help characterise the nature of the microstructural changes.

By contrasting these two regions, we want to assess the sensitivity of the proposed method. We scanned developing rats from postnatal day five (P5) to P30. We first used MEMRI to confirm the protracted cerebellar development compared to the thalamus. We then acquired diffusion properties of metabolites to probe short- and long-range restriction (i.e., soma restriction and branching). To interpret our data, we applied two biophysical models (one analytical and one computational) to characterise the specific changes in microstructure. We expected monotonic changes in microstructural parameters in both regions, reflecting cell growth, with thalamic properties plateauing earlier than cerebellar properties. This is mostly what we report in our observations. However, we report a U-shape trend for the cerebellar segment length (Lsegment, the distance between two embranchments of a dendritic tree) that matches openly available 3D cell reconstructions. In contrast to common findings in the adult brain, we report that tNAA and Glu do not provide reliable morphological neuronal markers in neonates. We identify instead total creatine (tCr) and taurine (Tau) as more reliable markers for tracking early neuronal development. We believe that this is a promising method to track early neuronal development, particularly in the cerebellum, where microstructure changes drastically postnatally.

Results

Structural imaging confirms protracted cerebellar growth

Representative slices from intra-age non-linear registration are shown in Figure 2A, underlying the cerebellar unfolding. Both cerebellar and thalamic volumes are just below 50µl at P5 (Figure 2B). However, the segmentation at P5 is poor, as explained in Methods (Figure S1), hence data at this time point should be treated cautiously, although the cerebellum and the thalamus, which are medial line structures, are less affected by the atlas distortions.

Manganese-enhanced MRI confirms protracted cerebellar growth in rat pups.

(A) Representative slices around the thalamus and cerebellum of non-linearly averaged images acquired at P5, P15 and P30. (B) Absolute volumetric changes are shown for the segmented thalamus and cerebellum. Large dots correspond to the mean and small dots to individual data points. (C) Normalised volumetric changes. Volume at each age is normalised by the P30 volume. Large dots correspond to the mean and small dots to individual data points.

The cerebellum triples in volume, between P5 and P30, reaching 90% of its P30 volume at P20, while the thalamus reaches 90% of the P30 volume at P15.

Metabolic profile changes with age

The metabolic profile changes noticeably between P5 and P30 as shown in Figure 3A. The absolute concentrations of metabolites, estimated by LCModel using the b0 averaged spectra, shown in Figure 3A, were normalised by the absolute concentration of macromolecules, MM (Figure 3B). Normalising by the MM content is an unusual choice. Conventionally, metabolite absolute concentrations are normalised by the water content in the voxel or by the tCr absolute concentration, which is considered stable in the healthy brain. However, this study compares regions with different relaxation properties at different ages, with different metabolites profiles in adulthood. The water relaxation properties are different between the thalamus and cerebellum, and both water relaxation properties and content change with age34,35. The tCr concentration changes substantially with age (about +50% in the cerebrum between P7 and P28) and is known to be higher in the cerebellum in adulthood (about +40-50% compared to the cerebrum)35,36. The MM concentration increases by about 30% between P7 and P28. MM resonances mainly come from the cytosolic proteins and although their content might slightly vary between regions in the healthy brain, they make a reasonable intracellular reference if SNR allows. We report normalisation by water in Figure S2, but the results are noisier than those normalised by MM. Between P5 and P30, metabolite/MM ratios increase for tNAA, Glu, Gln and Ins, and decrease for Tau, Gly, PE and tCho. Note that the increase in tNAA signal with age is protracted in the cerebellum compared to thalamus and has not reached a plateau at P30. These results generally match the metabolic concentrations measured with gold-standard classical MRS in the striatum, cortex and hippocampus of developing rats reported in Reference 35. We observe that tCr/MM increases strongly in the cerebellum between P5 and P30 but appears stable in the thalamus, while Reference 35 reported a 30% increase between P7 and P28 in the 3 ROIs. However, they also reported a 30% increase of the macromolecular content during this developmental period, matching our tCr/MM observation in the thalamus.

Metabolic profile changes postnatally.

(A) Spectra were binned per time point, region and b-value and averaged after phase and frequency correction. The SNR is clearly lower at P5/P10 in both regions: the voxel is about half of the P30 voxel size, and more animals were outliers. The SNR is also lower at P30: the rat size was slightly too big for the mouse cryoprobe at this age, so the skull is not at the top of the probe and the sensitivity is suboptimal (B) Spectra at b=0.035 ms/µm2 displayed in (A) were quantified with LCModel and absolute metabolite concentrations were normalised by the absolute macromolecular concentration and levelled to get Signal(tCr,P30)/Signal(MM,P30)=8 in the thalamus. See Figure S2 for a comparison with normalisation by water and individual estimations (rather than average spectra quantification).

Metabolite diffusion properties differ between regions and change with age

Thalamic (Figure 4A) and cerebellar (Figure 4B) signal attenuations up to b=30 ms/µm2 and ADCs at TM=100-1000 ms are shown for tNAA, Glu, tCr, Tau, tCho and Ins. These diffusion properties differ between both regions at each age (see Figure S3 for a region-to-region visual comparison). For most metabolites (except tCho), the signal attenuation is stronger in the thalamus, suggesting less restricted diffusion compared to the cerebellum. There is also a trend for lower signal attenuation and lower ADCs with increasing age for some metabolites, suggesting increasing restriction of metabolite diffusion. This is particularly striking for Tau in the thalamus. To give an estimation of the MRS fit performance, CRLBs are given for tNAA, Glu, tCr, Tau, tCho and Ins for the different diffusion conditions in Table S1.

Biophysical modelling issued from metabolites diffusion properties shows differential developmental trajectories in the cerebellum (A) and in the thalamus (B).

The first column (A,B) reports the signal attenuation as a function of b-value at TM=100ms, and the second column reports the ADC varying with diffusion time. Shadowed error bars represent the standard deviations. Modeling results from the spheres+”astrosticks” model is reported in the third column (A,B). P5 is shaded in the cerebellum due to possible motion artefact. The parameter space (sphere fraction, sphere radius) is displayed. Error bars represent the standard deviations. Llength, modeling output from the morphometric model is shown in the fourth column (A,B). P5 is shaded in the cerebellum due to possible motion artefact.

Biophysical modelling underlines different developmental trajectories of cell microstructure between the cerebellum and the thalamus

Data were fitted with a spheres + “astrosticks” model and a morphometric model (more details in the Methods section). The spheres + “astrosticks” model captures diffusion within spheres (representing cell bodies) and randomly oriented sticks (representing cell processes/neurites): parameters extracted are sphere fraction, fsphere (or equivalently neurite fraction as 1-fsphere) and cell radius, R. The parameter space (sphere radius vs sphere fraction) is represented in the 3rd column of Figure 4A (cerebellum) and Figure 4B (thalamus) for tNAA, Glu, tCr, Tau, tCho and Ins at all ages (see Figure S4 for a region-to-region visual comparison).

Mean values and standard deviations are reported in Table 1, along with the results from the linear mixed effect model ~ Age + Region + Age:Region + (1|PupID). The fits on averaged data, for binned spectra by age and region, are shown in Figure S5.

The morphometric model, only applied to averaged ADCs at long td, interprets diffusion in branched structures: the segment length, Lsegment is reported on the last column of Figure 4A/B (see Figure S6 for a region-to-region visual comparison and Figure S7 for all parameters estimated by the morphometric model: Dintra, NBranch, Lsegment).

Neuronal markers

tNAA and Glu are primarily concentrated in neurons in adulthood.

Thalamus: The sphere fraction and the radius estimated from tNAA and Glu diffusion properties do not change much with age. For both metabolites, fsphere is around 0.25 and R decreases to 5-6 µm. Lsegment is stable with age for tNAA, but shows a reduction at P15-P20 for Glu.

Cerebellum: The sphere fraction and the radius estimated from tNAA diffusion properties vary with age. fsphere nears 0.5 in the cerebellum and R decreases below 5 µm for tNAA and Glu. Lsegment decreases until P15 for both tNAA and Glu.

Between regions, sphere fraction differs significantly (Table 1).

Glial markers

tCho and Ins are primarily concentrated in glial cells in adulthood. The general trend is a high sphere fraction (~0.7-0.9) for Ins and moderately high (~0.5-0.6) for tCho. The radius estimated from tCho differs significantly between regions. Note the trend of decreasing sphere fraction in the cerebellum with age. Data at P15 for Ins in the thalamus were individually not reliable enough to apply the spheres + astrosticks model (c.f. Table S1 and the particularly higher CRLBs at P15 for Ins).

Non-specific markers

tCr and Tau are not known to be compartmentalised in a specific cell-type.

Thalamus: Parameter estimates for tCr diffusion properties do not change substantially with age (fsphere~0.35, R~6.5µm). Tau parameters estimation varies a lot with age: R shows a substantial decrease and fsphere increases.

Cerebellum: fsphere decreases from 0.63 (P10) to 0.41 (P30) for tCr, but R is stable. Tau follows a similar pattern, although with an overall lower fsphere (0.43 at P10 to 0.29 at P30). Lsegment increases in the cerebellum from P10 for tCr and increases P15 for Tau.

Between regions, the sphere fraction differs significantly for both metabolites (Table 1), but only Tau exhibits a significant change with age, and this change is different between regions (significant interaction term). For Tau, the radius is significantly different between regions, changes significantly with age and this change is different between regions (significant interaction term).

Discussion

As in Reference 2 we used MEMRI to validate the protracted development of the cerebellum in rat neonates3, unfolding and tripling volume between P5 and P30. Early developmental cell-specific cerebellar and thalamic microstructural features were assessed with dMRS.

We specifically report diffusion properties for neuronal markers tNAA and Glu, glial markers tCho and Ins, and the non-specific markers tCr and Tau. Metabolites exhibit significantly different diffusion properties between regions, demonstrating the relevance of dMRS to probing early cell structure development. fsphere is notably higher in the cerebellum for metabolites compartmentalised fully (tNAA, Glu) or partially (tCr, Tau) in neurons, suggesting the influence of the cell-body dense granular layer of the cerebellum. fsphere possibly reflects the relative volume of sphere-like structures (e.g., cell bodies) compared to the relative volume of fibrous structures (e.g., cell processes, fibres).

A study of human WM using diffusion MRI reported an increase in FA and a decrease in MD with age in early life.22 In mice, the same relationship was found in both WM and GM37. We report similar dMRI trends, in both cerebellum and thalamus (Figure S8).

We also report dMRS changes with age in both regions. The general trend we find for lower signal attenuation and ADC decrease (for some metabolites) with increasing age aligns with these reported dMRI observations, and could reflect overall microstructural complexification. However, the different nature of the signal (metabolites: mostly intracellular and low compartmental exchange vs. water: ubiquitous and fast exchange) prevents close co-interpretation.

fsphere(tCr, Tau) decreases with age in the cerebellum, meaning that tCr and Tau are located in compartments where the relative contribution of sphere-like structures (e.g. cell-bodies) decreases. In other words, it reflects cerebellar cell growth and complexification (Figure 4A, Figure 5A). In the cerebellum, while soma morphologic maturation mostly happens in the first week38, Purkinje cell dendritic expansion is rapid before P15 but then continues to mature up to the end of the 4th week38,39. Tau diffusion properties might reflect the development of Purkinje cells: (i) in cats, which cannot synthesise Tau and must source it from their diet, Tau dietary deficiency delays cortical and cerebellar maturation43, meaning that Tau is also crucial in early development, and (ii) in the cerebellum, immunohistochemistry studies locate Tau in adult Purkinje cells40–42. These observations align well too with the high Tau levels we still detect in the cerebellum at P30 (Figure 3B).

The biophysical parameters are interpreted based on the cerebellar microstructural development and main results are summarised.

Italic font corresponds to possible interpretations. (A) Schematic of cerebellar neuronal development and biophysical modelling parameters interpretation. Lsegment is interpreted as the distance between two branches of the dendritic tree, NBranch as the number of embranchments. The sphere radius Rsphere is interpreted as the average soma radius (i.e. granule and Purkinje cells). fsphere,LT represents the ratio of cerebellar layer thicknesses between the layers rich in sphere-like cell components (e.g., somas in the I/EGL, GL and PL) and the layers rich in fibrous cell components (e.g., cell processes or “neurite” in the ML). fsphere,LT can be seen as a rough approximation of fsphere and is inversely related to the dendritic tree growth, see Table S2. (B) Summary of the cerebellar results coming from the modelling. (1) and (2) refer to the models depicted in Figure 1. (C) Tau seems to change compartments in the thalamus with age, going from neuronal-like compartments (low sphere fraction) to glial-like compartments (high sphere fraction), also indicated by its important signal drop with age. It suggests that Tau is a marker of neuronal maturation. Its concentration remains high up to P30 in the cerebellum. Tau properties in the cerebellum (decreasing sphere fraction + relatively low sphere fractions) and contrast with the thalamus suggest that Tau could be a marker of cerebellar neuronal development.

The diffusion properties of glial metabolites, fsphere(Ins) and fsphere(tCho), also tend to decrease with age. This could be explained by another cerebellar peculiarity: Bergmann glia develop long processes along Purkinje cells and this specific growth could contribute to the fsphere(tCr,Tau) decrease.

Validation would require an atlas of all types of cerebellar cellular processes and bodies with estimates of their relative contribution. Such a resource would be immensely valuable but does not exist. Using literature data, we use the relative thicknesses of the EGL, IGL, PL (“cell body-like” layers) and ML (“fibre-like” layer) to derive fsphere,LT, a proxy of fsphere, as illustrated in Figure 5A and reported in Table S2 and Figure S11. fsphere,LT represents the contribution of sphere-rich layers compared to fibre-rich layers. The evolution of fsphere,LT trend with age is comparable to fsphere(tCr). Although fsphere(tCr) and fsphere(Tau) have similar trends, fsphere(Tau) values are overall lower, and below fsphere,LT values (Figure S11). This lends weight to the idea that Tau is preferentially located in cerebellar neurons during cerebellar development (and potentially less so in granule cells). In comparison, tCr, which appears more evenly spread between all cerebellar cell types, tracks the evolution of fsphere,LT more closely. Note that Tau has a higher Dintra44 compared to the other metabolites, potentially affecting the parameter estimation given for Tau. As described in Methods, the fit for Tau was substantially improved for Dintra > 0.6 µm2/ms. The results for Dintra=0.7 µm2/ms are similar for all the other metabolites (Figure S9), and generally improved the estimated sphere radius standard deviation. Additionally, the evolution with age of (R,fsphere) is then very similar for tNAA and Tau, reinforcing the idea that Tau is neuronal in the cerebellum.

In the thalamus, the decrease in Tau content (Figure 3B) and the change in Tau diffusion properties (Figure 4B) point towards a change of compartmentation with age. Contrary to the trend in the cerebellum, fsphere(Tau) consistently increases with age in the thalamus (Figure S9). This means that Tau diffuses in a space with fewer and fewer processes during development, which is incompatible with the hypothesis that Tau stays in the same compartment. Neurons, but also microglia45 or astrocytes46–48 develop with increasing ramification. Neurons are already branched at P10 in the thalamus33 (shown by a low fsphere(tNAA, Glu) at P10). fsphere(Tau) at P10 nears fsphere(tNAA, Glu), indicating that Tau might be located in neuronal compartments in the neonatal thalamus. Although this hypothesis is not verifiable in existing literature, some in vitro studies show that Tau contributes to neuronal proliferation49 and neuronal growth50. At P30, fsphere(Tau) is closer to fsphere(tCho,Ins), indicating that Tau might be located in glial compartments in adulthood. Indeed, immunohistochemistry studies locate Tau in adult thalamic glia42.

We suggest Tau as a cerebellar marker of early neuronal development, possibly more specifically of Purkinje cells. We confirm in this study the observation from Reference 35 that both tCr and Tau have high spectroscopic signals in neonates. They are the two metabolites most reliably quantified up to P15 in both regions (continuing for Tau in the cerebellum up to P30), and the estimates of their diffusion properties are the most confident, contrary to the usual neuronal markers, tNAA and Glu. tCr stands as a good marker, too. However, there is no indication of a preferential neuronal compartmentation for this metabolite during early development, hence it is less specific to neurons, and probably even less to Purkinje cells. Because Tau seems to switch compartments in the thalamus between birth and adulthood, we do not think its diffusion properties reliably inform thalamic neuronal morphological development. In both human51,52 and rhesus monkey53 brains, Tau content increases at birth and then decreases with age, and is therefore a marker of the immature brain in both species. Although Tau concentration is low in the adult human brain, it is more concentrated in the cerebellum in adulthood52,54. A similar finding was made in the adult rhesus monkey brain53, indicating a conserved pattern across species of Tau properties in the cerebellum. These observations form a good basis for translation. Although dMRS in infants has substantial challenges, they are worth trying to overcome for a putative cerebellar marker of early neuronal development which could help understand neurodevelopmental disorders 3.

We also observed a protracted increase in tNAA signal in the cerebellum compared to the thalamus. tNAA is usually described as a neuronal osmolyte, meaning it could reflect neuronal growth and maturation and be a marker of neuronal development. However, at the voxel level, its signal reflects the neuronal volumetric contribution, i.e. differences in neuronal densities might mask the effect of development. In contrast, we want to reiterate that dMRS is a normalised measure and dMRS outcomes are not affected by the neuronal density: they represent the average morphological properties of the cell types the metabolite belongs to.

We chose to interpret the dMRS data with two biophysical models: an analytical spheres & “astrosticks” model and a computational morphometric model. The analytical model was applied simultaneously to the signal attenuation at high b-values and the ADCs at long diffusion times, and on individual animals. In addition to the valuable biophysical biomarkers, it provides a good signal representation to evaluate the significance of the differences between diffusion properties of individual datasets. The experimental design was not suitable for a randomly oriented infinite cylinders (“astrocylinders”) model as in previous studies28,29. The mixing time used for the acquisition of signal decay at high b-values (TM=100 ms) is too long to enable sensitivity to fibre radius26, hence the choice of “astrosticks” model. The sphere compartment was added to account for the high sphere fraction in the cerebellum, and for the nature of the neonatal brain (under-developed neurons). The RMSEs from the different models tested (Figure S3) tend to be more similar at P30, suggesting that it is less important to add a sphere compartment for modelling metabolites diffusion in the adult brain.

The computational morphometric model was applied to attempt to extract cell extension-related parameters, but the study design and data quality are suboptimal, and we could only reliably extract Lsegment and Dintra (Figure S7). Dintra is mostly stable with age in both regions. However, it sometimes drops or peaks, likely affecting the other parameters’ estimations. The longest mixing time (TM=1000 ms) is short compared to previous studies using a similar modelling approach27,29, meaning that the diffusion path is shorter (i.e., about 30µm in this study compared to about 45µm in previous studies, assuming Dintra=0.5 µm2/ms). Note that the higher Dintra of Tau is therefore an asset for probing early neuronal cerebellar growth. In our study, the number of embranchments (Nbranch) was very poorly evaluated because of the high level of noise in the data. However, even with a better dataset quality and a more optimal study design, it is unlikely that the Nbranch estimation could reflect the Purkinje cell morphometry because of their size and complexity in adulthood. If we had robust DTI in all animals (Figure S8), we would have included the fibre dispersion in the modelling (notably for the cerebellum), as in Reference 29. However, since we acquired enough directions for the dMRS data to be powder-averaged, we expect that the WM contribution in the cerebellum is mostly reflected by a slight fsphere drop, as exemplified by the fsphere,LT drop when including WM (Table S2).

Finally, we expected that Lsegment estimated from the morphometric model in the cerebellum would increase regularly with age, matching the dendritic tree development. For neuronal, or partly neuronal metabolites, we report an Lsegment decrease first. Except for tNAA, which consistently decreases in the cerebellum with age, Lsegment decreases until P15 for tCr, Tau and Glu and then increases, forming a U-shape curve. To better understand these results, we extracted the morphometries of developing Purkinje cells and developing thalamic interneurons (both in the mouse) from NeuroMorph.org. Interestingly, from these 3D reconstructed cells, we observe a decrease in Lsegment, 3D-recon up to P13 for Purkinje cells, whilst being quite stable in the thalamus (Figure 6), matching our non-invasive measurement trend. This could reflect the complexification of the dendritic tree following its expansion.

Morphometric parameters extracted from mouse real cells 3D reconstructions on NeuroMorph shows a decrease of Lsegment up to P13 in developing cerebellar Purkinje cells, but not in developing thalamic interneurons.

(A) Examples of cells extracted from NeuroMorph. The total number of cells used per age and region is given in Supplementary Material. References: Cerebellum P6-P1055, P10-P1356, P35-P4357; Thalamus P9-P3058 (B) The values of Branch Length (Lsegment), Branching Order and total process length are directly estimated from the morphology using the function stat_tree in the Trees Matlab Toolbox.

Conclusion

This study proposes diffusion-weighted MRS as a method to longitudinally and non-invasively measure cell-specific cerebellar and thalamic microstructural features during early development in rat neonates. We demonstrate the sensitivity of Taurine to early neuronal cerebellar development. This is a stepping-stone enabling further studies in pathological conditions.

Methods

Animals

Experiments were approved by the United Kingdom Home Office. 18 Sprague-Dawley rats (9 females) were scanned at P5, P10, P15, P20 and P30. Pups came from 3 litters (6 pups scanned per litter). Animals were induced with 4% isoflurane in oxygen and anaesthesia was maintained with 1-2% isoflurane in oxygen. Respiration was monitored with a breathing pillow. Temperature was monitored with a rectal probe, which was used as a cutaneous probe from P5 to P20 to avoid any risk of harm. Isoflurane level was adapted to keep the animals at 50-70 rpm and a warm water-circulation heating pad was used to keep the animals at 36-37°C during scanning. Pups were gently held with foam and protected tape to minimise motion, because the neonatal skull is not suitable for the use of ear bars.

For logistic reasons we could not acquire MEMRI for all litters. Animals from the second litter had manganese-enhanced anatomical scans. To this end, they were intraperitoneally injected with a 50 mg/kg solution of MnCl2 24 hours prior to the scan. Under the age of P10, the dam received the intraperitoneal injection, and the manganese was transferred to the neonates via the maternal milk2.

MRI acquisitions

Data were acquired on a 7T Biospec Bruker MRI scanner (Paravision 360.1, max gradient strength: 660 mT/m), equipped with a mouse quadrature cryoprobe, well-suited for rat neonates. An anatomical scan around the region of interest (thalamus or cerebellum) to place the voxel (T2-weighted, 2D RARE, R=8, 0.2 mm isotropic resolution, TE=33 ms, TR=3500 ms). For 6 animals, this scan was replaced by a whole brain manganese-enhanced anatomical scan (T1-weighted, 3D MGE, TE=2.4 s, TR=62 ms, 0.15 mm isotropic resolution).

dMRS acquisitions

A STELASER59 sequence (TE=34 ms, diffusion gradient duration, δ=5 ms) was used for dMRS acquisitions. The voxel was placed in the cerebellum for 9 pups, and in the thalamus for the other 9 pups. Voxel size was increased with age to follow brain growth. Diffusion-weighted spectra were acquired at 4 mixing times (TM) and 2 b-values (c.f. Table 2 for detailed parameters): a small “b0” used as a crusher for the stimulated echo, and b0+3 ms/µm2. Higher b-values were also acquired at the shortest diffusion time to probe restriction within cell bodies and enable inference of cell fraction and radius. The macromolecular baseline was acquired in the adult rodent with the same STELASER parameters (TE and δ), at the shortest TM=100 ms, but the stimulated echo was preceded by a double-inversion recovery (TI1=2000 ms, TI2=710 ms). The b-value was set to 10 ms/µm2 to reduce the residual contribution of bulk metabolites, and the tCr peak residual at 4ppm (due to the shorter T1) was corrected by subtracting a Lorentzian.

Processing

dMRS

Processing was achieved with a custom Matlab (The Mathworks) routine (https://github.com/clemoune/dmrs-neonates/tree/main/A_dMRS_processing):

Signals from different coils were combined, using the water signal at the lowest b-value and shortest diffusion time.

Outliers were identified and discarded, as described in60.

For each transient i, the mean absolute upfield signal (range 1.9-4.5 ppm), Mup(i) was computed.

Transients affected by motion are characterised by a clear drop of Mup(i). Transients falling below the median(Mup) - 3xMAD threshold were discarded.

Experiments at high b-values are more susceptible to changing gradient direction (particularly if the voxel contains white matter tracts). To predict its effect across b-values and overcome the small number of acquisitions per unique condition (i.e. a (b-value, direction) pair), a diffusion kurtosis tensor was computed based on Mup. The MAD calculated at b=0 was used as a “ground truth” MAD and was reported to higher b-values (per direction). High b-values acquisitions are more easily corrupted by motion and therefore present an artificially high MAD.

Individual transients were corrected for phase and frequency drifts and averaged across all diffusion directions (powder-averaged) via an arithmetic mean. In general, metabolites signal was visually detectable on single transients (with a line-broadening of 5 Hz), except at b=20,30 ms/µm2 at P5, rendering these acquisitions more susceptible to motion artefacts.

Any water residual was removed using a singular value decomposition method centred around the water region. If the baseline was heavily distorted despite this step due to very poor water suppression, the acquisition was discarded from the analysis.

Spectra were subsequently fitted with LCModel. The simulated basis set contained the following metabolites: Ace, Ala, Asp, Cr, GABA, Gln, Glu, Gly, GPC, GSH, Ins, sIns, Lac, NAA, NAAG, PCho, PCr, Tau, PE, and included a macromolecules (MM) baseline acquired experimentally (double-inversion recovery and b=10 ms/µm2).

Diffusion properties of NAA+NAAG (tNAA), Glu, Cr+PCr (tCr), Tau, GPC+PCho (tCho) and Ins are reported. In order to model both signal compartmentalisation and restricted diffusion by cell membranes, we investigated metabolites diffusion through two sets of measurements: i) the direction-averaged signal decay as a function of the diffusion weighting b, keeping the diffusion time constant at 104.25 ms (time between diffusion gradients Δ fixed at 105.91 ms and δ fixed at 5 ms) and varying the strength of the diffusion-sensitizing gradients; and ii) the apparent diffusion coefficient (ADC) as a function of increasing diffusion time (keeping δ fixed at 5 ms). For the measurements i), data are reported as the ratio S(b)/S(b=0). For the measurements ii), ADCs diffusion time dependencies are reported. ADCs are calculated following . Negative ADCs were systematically removed as they point to a quantification error.

In the cerebellum, metabolites’ signal attenuation at P5 is always much stronger and quite far off the other time points. The macromolecules’ signal attenuation at high b-values is supposed to remain low59 and can be used as an internal probe for motion artefact44. However, the macromolecules’ signal attenuation is quite strong at P5 in the cerebellum (Figure S8), pointing out to a residual motion artefact despite careful processing. The spectroscopic signal in individual transients was too low at high b-values, notably in the cerebellum: it is possible that the spectral realignment failed.

MEMRI

The three first echoes of each MGE acquisition were averaged and denoised using the DenoiseImage function from the MINC toolbox, based on61. For each time point, an image with no obvious artefact and a good visual CNR was taken as the initial model and transformed into the Fischer atlas62 space. Images were then processed using registration_tamarack.py with the MAGeT segmentation in the available Pydpider pipelines63. Due to the large deformations between P30 and P5, the P30 timepoint could not be used as a reference. An adequate atlas from P15 was generated in a couple of iterations (using a P20 atlas as intermediary). It was then used as a reference atlas and reference timepoint for the pipeline. The olfactory bulb and the more frontal parts were poorly registered because the image often had an intensity drop there, but it did not affect the registration of caudal regions, such as the cerebellum and the thalamus. Image with strong intensity differences, or corrupted by motion were removed from the analysis. The registration did not work well at P5, even when attempting to use an atlas generated at P10 (that is already slightly distorted).

Analysis

Signal decays at high-b values (at TM=100 ms) and ADCs at varying diffusion times were fitted together with a spheres and “astrosticks” (corresponding to randomly oriented sticks) model using the functions implemented in DIVE (https://git.fmrib.ox.ac.uk/fsl/DIVE). The DIVE toolbox uses the Gaussian phase distribution approximation We fitted the signal to: Free parameters were R (sphere radius) and fsphere (sphere fraction). Dintra (i.e. the metabolite intracellular free diffusivity) was fixed for both compartments at 0.5 µm2/ms. Previous studies measuring brain metabolites diffusion properties at ultra-short diffusion-times44,64 estimated the intracellular free diffusivity at 0.5-0.6 µm2/ms for most metabolites (except Tau). However other studies using different (less direct) approaches reported lower Dintra27,28, hence we tested a range of reasonable Dintra. Dintra was fixed 0.3, 0.4, 0.5, 0.6, 0.7 and 0.8 µm2/ms. RMSEs between model fit and data were equivalent for all Dintra, except for Tau in the thalamus where increasing Dintra improved the fit, particularly until Dintra=0.5µm2/ms. Moreover, parameters estimation was unrealistic for low Dintra. Increasing Dintra between 0.5 µm2/ms and 0.7 µm2/ms did not change the trend of the results, except for Tau in the thalamus (see Figure S0 and Discussion). The fit was made robust to missing data (when excluded during processing).

To test for the model accounting best for our data, we tested different analytical models. RMSE between model fit and data were calculated for “astrosticks” only (1 free parameter, Dintra), “astrocylinders” (corresponding to randomly oriented infinite cylinders, 2 free parameters: Dintra and cylinder radius), spheres and astrosticks without constraint (3 free parameters) and spheres and astrosticks with Dintra fixed at 0.5 µm2/ms (Figure S0). RMSE were the lowest for spheres and astrosticks, and were very similar with and without the Dintra constraint. Hence the model with the least free parameters was chosen. Moreover, the results from the constrained spheres and astrosticks model were more stable and physically meaningful with the constraint on Dintra. The code used for the analysis can be found on: https://github.com/clemoune/dmrs-neonates/tree/main/B_Astrosticks_Spheres_model.

Long diffusion times

Average ADCs at varying diffusion times were fitted with the morphometric model introduced in27. Modeling was performed using the dictionary generated for29 with TM = 100, 750 ms predicted by interpolation, and a random forest regressor (‘treeBagger’ function in MatlabR2020b, with tree depth = 20; number of trees = 200 and other model hyper-parameters as per default) trained in a fully supervised fashion, using the ADCs time dependencies as input and the ground truth morphological parameters as targets; as loss function the mean squared error between predicted and ground truth parameters values was used. Parameters fitted were Dintra, Nbranch, Lsegment, SDNbranch and SDLsegment. Boundaries for fitting the model to data were respectively [0.1, 1], [2, 25], [5, 100], [2, 3] and [5, 10]. Given the data noise, only Lsegment and Dintra are estimated robustly. Fits for all parameters are shown in Supplementary Information.

Statistical analyses

Statistical analyses were run with R, using the “lmerTest65 package. To assess whether diffusion properties were significantly different in the 2 regions during development, we applied a linear mixed effect model to the parameters extracted from the spheres + “astrosticks” model (sphere fraction and sphere radius), using age and region as fixed effects, and a random intercept for pups ID. An analysis of variance enabled to identify whether the inclusion of an interaction term age*region was necessary or not. Results were corrected for multiple hypotheses testing with a Bonferroni correction (6 metabolites,2 parameters = 12 hypotheses). The threshold corrected p-value is 4.2e-3 for α=0.05 and 8.3e-4 for α=0.01.

eLife Assessment

This study presents a valuable investigation into cell-specific microstructural development in the neonatal rat brain using diffusion-weighted magnetic resonance spectroscopy. The evidence supporting the core claims is solid, with innovative in vivo data acquisition and modeling, although some conclusions would benefit from stronger validation and methodological justification. The work will be of interest to researchers studying brain development and biophysical imaging methods.

Reviewer #1 (Public review):

In this work, Ligneul and coauthors implemented diffusion-weighted MRS in young rats to follow longitudinally and in vivo the microstructural changes occurring during brain development. Diffusion-weighted MRS is here instrumental in assessing microstructure in a cell-specific manner, as opposed to the claimed gold-standard (manganese-enhanced MRI) that can only probe changes in brain volume. Differential microstructure and complexification of the cerebellum and the thalamus during rat brain development were observed non-invasively. In particular, lower metabolite ADC with increasing age were measured in both brain regions, reflecting increasing cellular restriction with brain maturation. Higher sphere (representing cell bodies) fraction for neuronal metabolites (total NAA, glutamate) and total creatine and taurine in the cerebellum compared to the thalamus were estimated, reflecting the unique structure of the cerebellar granular layer with a high density of cell bodies. Decreasing sphere fraction with age was observed in the cerebellum, reflecting the development of the dendritic tree of Purkinje cells and Bergmann glia. From morphometric analyses, the authors could probe non-monotonic branching evolution in the cerebellum, matching 3D representations of Purkinje cells expansion and complexification with age. Finally, the authors highlighted taurine as a potential new marker of cerebellar development.

From a technical standpoint, this work clearly demonstrates the potential of diffusion-weighted MRS at probing microstructure changes of the developing brain non-invasively, paving the way for its application in pathological cases. Ligneul and coauthors also show that diffusion-weighted MRS acquisitions in neonates are feasible, despite the known technical challenges of such measurements, even in adult rats. They also provide all necessary resources to reproduce and build upon their work, which is highly valuable for the community.

From a biological standpoint, claims are well supported by the microstructure parameters derived from advanced biophysical modelling of the diffusion MRS data.

Specific strengths:

(1) The interpretation of dMRS data in terms of cell-specific microstructure through advanced biophysical modelling (e.g. the sphere fraction, modelling the fraction of cell bodies versus neuronal or astrocytic processes) is a strong asset of the study, going beyond the more commonly used signal representation metrics such as the apparent diffusion coefficient, which lacks specificity to biological phenomena.

(2) The fairly good data quality despite the complexity of the experimental framework should be praised: diffusion-weighted MRS was acquired in two brain regions (although not in the same animals) and longitudinally, in neonates, including data at high b-values and multiple diffusion times, which altogether constitutes a large-scale dataset of high value for the diffusion-weighted MRS community.

(3) The authors have shared publicly data and codes used for processing and fitting, which will allow one to reproduce or extend the scope of this work to disease populations, and which goes in line with the current effort of the MR(S) community for data sharing.

Specific weaknesses:

Ligneul and coauthors have convincingly addressed and included my comments in their revised manuscript.

I believe the following conceptual concerns, which are inherent to the nature of the study and do not require further adjustments of the manuscript, remain:

(1) Metabolite compartmentation in one cell type or the other has often been challenged and is currently impossible to validate in vivo. Here, Ligneul and coauthors did not use this assumption a priori and supported their claims also with non-MR literature (eg. for Taurine), but the interpretation of results in that direction should be made with care.

(2) Longitudinal MR studies of the developing brain make it difficult to extract parameters with an "absolute" meaning. Indirect assumptions used to derive such parameters may change with age and become confounding factors (brain structure, cell distribution, concentrations normalizing metabolites (here macromolecules), relaxation times...). While findings of the manuscript are convincing and supported with literature, the true underlying nature of such changes might be difficult to access.

(3) Diffusion MRI in addition to diffusion MRS would have been complementary and beneficial to validate some of the signal contributions, but was unfeasible in the time constraints of experiments on young animals.

Reviewer #2 (Public review):

This second revision has partially addressed criticisms previously raised; however, substantial inadequacies, particularly concerning rigorous validation and model justification, remain unresolved. While recognizing evident strength, novelty, and technical complexity of this work, the authors have yet to fully resolve key major concerns explicitly pointed out during revision in a satisfactory manner. As currently written, the manuscript does not yet provide sufficiently robust validation, methodological rigour, or clarity required for complete acceptance in a top-tier scientific journal.

Summary of Authors' Aim:

In this revised version, the authors aimed to address prior reviewer critiques harshly pinpointing the need for greater clarity in the manuscript's logical flow, rigorous external validation, clearer explanation of methodological normalization choices, and deeper elaboration of diffusion MRI method relevance and potential translation. The authors present a diffusion-weighted MRS approach paired with complex biophysical modelling to elucidate differential developmental trajectories of cellular structures in cerebellum and thalamus in rat neonates, providing a novel, non-invasive avenue for monitoring cellular microstructure.

Major Comments:

Rigorous Validation (Reviewer #1 - point R1.1, Reviewer #2 - point R2.2):

The major concern previously raised and reiterated here is the insufficient external cross-validation of the dMRS-derived interpretations about cellular changes, including the particularly speculative interpretation that taurine undergoes compartment switching between neuronal and glial compartments in the thalamus. The authors acknowledge this important shortcoming (R1.1, R2.2) but attempt to mitigate these concerns merely through additional contextual comparisons from existing literature (page 23, lines 877-878, Figure S11, Table S2). While better contextualization is welcome, the modified manuscript still falls notably short of the level of rigour necessary to validate such striking switches in compartmentalization. To justify claims of metabolites changing cellular compartments, explicit verification against independent molecular/histological data, ideally with additional immunohistochemical staining for cellular markers (e.g., glial fibrillary acidic protein, NeuN), is necessary. The mere presence of literature correlations (such as the reported visual comparisons to morphometric reconstructions, page 24, lines 883-884) does not constitute rigorous validation at the required standard for high-impact publication. The revised manuscript remains fundamentally weakened without such validation. To properly improve, the authors must consider incorporating independent ex vivo experiments or, if this is no longer feasible, extensively temper their compartment-switching claims, acknowledging explicitly and prominently the speculative nature of current interpretations.

Normalization of Metabolite Concentrations (Reviewer #1 - point R1.3):

The authors clearly responded to a reviewer wish for justification of metabolite normalisation to macromolecular concentrations (page 13, lines 493-503, Figure S2). However, the rationale provided remains only partially convincing. While the authors appropriately acknowledge the unusual nature of their methodological choice and possible confounding factors, they opt to supplement rather than substitute this approach with a more standard method (normalisation by water) in the main body of the manuscript. The additional supplementary Figure S2 is helpful, yet the conclusions derived with macromolecular normalization still remain potentially confounded by age-dependent macromolecular changes (Tkac et al., 2003). The justification given in the revised manuscript remains vague, unsatisfactory, and somewhat contradictory-authors accept macromolecules changes likely with age, yet largely overlook this effect. At least, the comparison between normalization by macromolecules and water should be explicitly discussed in the main text, and conclusions drawn from macromolecular normalization must be cautiously framed.

Choice and Justification of Biophysical Model (Reviewer #1 - point R1.4):

The reviewers questioned model assumptions, particularly ignoring macroscopic anisotropy effects due to white matter presence, myelination, and fibre orientation dispersion in the cerebellar voxel. Authors provided newly included DTI data and acknowledged this limitation explicitly (R1.4, Figure S8, page 25, lines 921-924). However, the addition of these poor-quality DTI data with limited interpretability paradoxically weakens rather than strengthens the manuscript as a whole, since the authors now present unclear supplementary results with little additional interpretative value. Recognizing poor data quality in this scenario, although intellectually honest, does not substantially increase the current robustness of their chosen model nor improve justification. To address this fully, either higher-quality data should be collected to robustly probe anisotropy or fibre dispersion effects, or the authors must much further restrict their interpretations in view of this clear limitation. Currently, the solution proposed is incomplete and insufficient to clarify the consequences of their chosen model.

Logical Flow and Clarity (Reviewer #2 - points R2.1 and R2.3):

The authors attempted to respond to reviewer comments on logical flow and accessibility (page 3, introduction restructuring). While the manuscript readability has improved, the introduction and discussion remain overly intricate, and at times, detail-oriented without clear links into central claims. In particular, the biological rationale for choosing the specific metabolite markers (especially tCho, Ins, Tau, etc.) and their known relevance must be further streamlined and simplified to increase accessibility and directness. Although some helpful restructuring was carried out, further careful paragraph-level revision for logical flow and readability remains necessary.

Translation to Human Studies (Reviewer #2 - point R2.4):

The authors have extended contextual discussion on translational potential regarding taurine as a developmental marker in humans (pages 24-25, lines 906-917). However, mention remains vague and cursory, without presenting sufficiently solid arguments nor drawing from human developmental studies adequately. Translational potential must be assessed within the realistic limitations inherent in clinical translation of MRS studies, particularly given the technical complexities clearly identified even in preclinical studies of this paper. Discussion remains relatively superficial, and if retained, must be expanded to fully discuss realistic human translational hurdles and requirements.

Author response:

The following is the authors’ response to the previous reviews

We thank the reviewers once again for their careful evaluation of the revised manuscript and for their constructive suggestions. In response to the remaining recommendations, we have made minor amendments to the manuscript. The main changes are as follows:

• Metabolite Concentrations: we now report them more conventionally, i.e. normalised by water content. The original normalisation by the absolute MM content has been retained in the supplementary information, as MMs are an endogenous tissue probe (i.e., not dependent on cerebrospinal fluid). The fact that both water and MM normalisation provide similar trends supports the robustness of our conclusions. We have also updated Figure S2 to include the absolute MM concentrations, raw water content, and the MM-to-water ratios for each time point.

• Taurine Interpretation: We have revised the wording related to the interpretation of taurine findings to clarify that we present a set of converging observations suggesting taurine may serve as a marker of early cerebellar neurodevelopment, rather than asserting it as a definitive conclusion.

Comments to the editor & reviewers:

We sincerely thank the reviewers and the editor for their valuable feedback, which has significantly improved the manuscript since its initial submission.

Please note a correction in Figure S2 (added during the previous revision round): the reported evolution of metabolite/water concentrations has changed due to an earlier error in calculating the water peak integral, which has now been corrected.

While we recognise that a study and manuscript can always be improved, we prefer not to make further changes at this stage. We cannot conduct new experiments, and redesigning the model falls outside the scope of this work. Additionally, we believe that further altering the manuscript’s structure could lead to unnecessary confusion rather than clarity.

Version 3

Abstract

Abstract

It is currently impossible to non-invasively assess cerebellar cell structure during early development. Here we propose a novel approach to non-invasively and longitudinally track cell-specific development using diffusion-weighted magnetic resonance spectroscopy in combination with microstructural modelling. Tracking metabolite diffusion allows us to probe cell-specific developmental trajectories in the cerebellum and thalamus of healthy rat neonates from post-natal day (P) 5 to P30. Additionally, by comparing different analytical and biophysical microstructural models we can follow the differential contribution of cell bodies and neurites during development. The thalamus serves as a control region to assess the sensitivity of our method to microstructural differences between the regions. We found significant differences between cerebellar and thalamic metabolites diffusion properties. For most metabolites, the signal attenuation is stronger in the thalamus, suggesting less restricted diffusion compared to the cerebellum. There is also a trend for lower signal attenuation and lower ADCs with increasing age, suggesting increasing restriction of metabolite diffusion. This is particularly striking for taurine in the thalamus. We use biophysical modelling to interpret these differences. We report a decreased sphere fraction (or an increased neurite fraction) with age for taurine and total creatine in the cerebellum, marking dendritic growth. Surprisingly, we also report a U-shape trend for segment length (the distance between two embranchments in a dendritic tree) in the cerebellum agreeing with age-matching morphometry of openly available 3D-Purkinje reconstructions. Results demonstrate that diffusion-weighted MRS probes early cerebellar neuronal development non-invasively.

Introduction

The cerebellum is a well-conserved structure across species and particularly across mammals. The cerebellar cortex is a highly folded structure organized in three layers: the molecular layer (containing the notable Purkinje cell dendritic trees), the Purkinje layer (containing the somas of Purkinje cells), and the granular layer (containing the granule cells). Projections to the deep cerebellar nuclei form the cerebellar white matter. Despite its small size (compared to the cerebrum), it contains more than half of the total brain neurons: this is mainly due to the granule cells, very small and densely packed cells, some of the smallest of the brain, presenting a very simple morphology1. On the other hand, the Purkinje cells are one of the largest types of neurons, and most complex cells in the brain. Cerebellar development is protracted2, starting from birth (or the last trimester of pregnancy) to a few weeks old in rodents (or 2 y.o. in humans)3. The cerebellum is involved in many basic and high-level functions, from motor control to social skills, and it is known to be affected in a number of neurodevelopmental disorders4.

During early development, neurons arborise. Their morphology becomes more complex and exhibits denser branching. In rodents, Purkinje cells’ dendritic trees start growing at birth. In 30 days, they evolve from a simple soma (∼10 µm diameter) to a large cell with a highly branched and planar dendritic tree (soma diameter ∼20 µm, area ∼10,000 µm2 in the mouse5). Purkinje cell morphology can be altered in autism spectrum disorders (ASD), as shown post-mortem, in humans6 and multiple mouse models of ASD. Specifically, they show an increased spine density in Tsc1 mutant mice7, an overgrowth of the dendritic tree and abnormal branching in PTEN KO mice8, a reduction of the dendritic tree in AUTS2 mice9; and general alterations in developmental disorders10–13.

All cerebellar outputs transit via the deep cerebellar nuclei (DCN), and a couple of them project to the thalamus before spreading to the cortex. The thalamus is considered a relay station in the brain and contains about 60 nuclei on unique pathways. Some thalamic nuclei are involved in psychiatric disorders, and more specifically, the microstructure of the mediodorsal nucleus and the pulvinar nucleus are altered in schizophrenia and psychosis14. Some other nuclei are involved in neurodevelopmental disorders (e.g., the lateral reticular nuclei in ASD15). Generally, these disorders are associated with an altered thalamic connectivity14. Thalamocortical projections are crucial for the cytoarchitectural development of the cortex16,17.

Neurodevelopmental disorders, or even psychiatric disorders with a neurodevelopmental origin such as schizophrenia18, generally need better neurobiological characterisation during development19.

Rodents reach adulthood after a couple of months and are an ideal model for mammal brain development. While brain cell developmental trajectories are key to assessing neurodevelopment, there is a lack of translational techniques allowing for a non-invasive measure of their properties. Typically, EEG and MEG signals contain brain activity cell-specific information. Still, it is hard to retrieve in practice, and the existing forward or inverse models notably use prior information on cell morphology. PET can be cell-specific and provide information about metabolism, but it is blind to cell morphology and relies on radioactive ligands, making it very expensive and potentially harmful. fNIRS and ultrasounds are too indirect to measure cell-specific information as they rely on the hemodynamic response. Last but not least, MRI is non-invasive and non-harmful. The MR signal can be sensitised to assess different tissue properties. In the following paragraphs, we will summarise the importance and limitations of some MR methods sensitive to brain structure and microstructure during development and introduce our primary method, namely diffusion-weighted MR spectroscopy (dMRS).

(Structural MRI: anatomy) Manganese-enhanced MRI (MEMRI) is particularly suited to studying whole-brain early development in rodents. It provides high structural contrast and a fast acquisition time in neonates2,20. Still, the method is blind to the microstructural changes that lead to the volumetric changes reported.

(dMRI: microstructure) Diffusion-weighted MRI (dMRI) measures the diffusion properties of water in the brain: restricted and hindered by cell membranes, water diffusion provides an indirect measure of the tissue microstructure21. dMRI is versatile and applied in numerous applications, including the characterisation of the microstructural organisation during neurodevelopment in rodents and, to a lesser extent, in humans. Comparison with postmortem microscopy shows a correlation between a high cortical fractional anisotropy (FA) at birth and the presence of radial glia. The latter progressively disappears in the first week along with an FA decrease, illustrating the potential of dMRI to characterise neurodevelopment22,23. In the cerebellum, the dense folding limits diffusion imaging to exploring white matter or ex vivo samples24.

(dMRS: cell-specific microstructure) Water is ubiquitous and can only partially disentangle the contribution from different tissue compartments (e.g. intra and extracellular space). Diffusion- weighted magnetic resonance spectroscopy (dMRS) measures the diffusion properties of metabolites. Brain metabolites are mostly intracellular, with some more specific to neurons (e.g., N-acetyl- aspartate, NAA, and glutamate, Glu) and others more specific to glial cells (e.g., choline compounds, tCho, and myo-inositol, Ins), therefore different cellular compartments can be resolved25. However, metabolites are 104-fold less concentrated than water, requiring large measurement volumes for dMRS (typically ∼ 50-100mm3 in rodents), and long acquisition times. As for dMRI, some dMRS acquisition parameters, such as the diffusion weighting and diffusion time, enable the signal to be sensitised to different cellular morphological scales (e.g., cellular processes, cell bodies, cell extension). Using multiple diffusion times and a post-hoc model, cellular morphological parameters can be estimated from dMRS26. They correlate with neuronal and glial morphologies extracted from histological measures in the healthy, adult mouse27,28. More specifically, it has been shown that myo- inositol diffusion properties reflect astrocytic morphology in an induced model of astrocytic hypertrophy29, and in a mouse model of demyelination where astrocytic activation was expected30. Choline compound’s diffusion properties were associated with microglial activation in human disease31, or in an LPS-induced brain inflammation model32. However, to our knowledge, very few studies have used dMRS to probe changes in neuronal morphology, and none have attempted to probe neuronal development in early life.

Here, we propose a non-invasive method to monitor cellular brain development. We focus on two regions with different developmental timelines: the cerebellum and the thalamus. As described above, the cerebellum expands greatly after birth, when the Purkinje cells dendritic trees start growing. This slow maturation exposes the cerebellum longer to developmental insults. In contrast to the cerebellum, the thalamus develops relatively quickly, and its neurons expand and complexify more rapidly2,33 (Figure 1A).

Overview of the rationale of the study.

(A) Context: the cerebellum exhibits protracted growth compared to other brain regions (e.g., thalamus), making it a particularly vulnerable region for neurodevelopmental disorders. Its specific microstructural organisation (i.e. layered with Purkinje cells and granular cells) changes between birth and adulthood. (B) The current “standard” is to probe regional growth with manganese-enhanced MRI (MEMRI). However, it is not sensitive to the underlying microstructural changes. (C) Pushing limits: diffusion-weighted MRS (dMRS) can potentially assess cell-specific microstructural changes in given regions during development. Different measures of metabolites diffusion properties (apparent diffusion coefficient, ADC, at long diffusion times, signal attenuation, S, at high diffusion weighting, b-values) can be interpreted with biophysical modelling and help characterise the nature of the microstructural changes.

By contrasting these two regions, we want to assess the sensitivity of the proposed method. We scanned developing rats from postnatal day five (P5) to P30. We first used MEMRI to confirm the protracted cerebellar development compared to the thalamus. We then acquired diffusion properties of metabolites to probe short- and long-range restriction (i.e., soma restriction and branching). To interpret our data, we applied two biophysical models (one analytical and one computational) to characterise the specific changes in microstructure. We expected monotonic changes in microstructural parameters in both regions, reflecting cell growth, with thalamic properties plateauing earlier than cerebellar properties. This is mostly what we report in our observations. However, we report a U-shape trend for the cerebellar segment length (Lsegment, the distance between two embranchments of a dendritic tree) that matches openly available 3D cell reconstructions. In contrast to common findings in the adult brain, we report that tNAA and Glu do not provide reliable morphological neuronal markers in neonates. We identify instead total creatine (tCr) and taurine (Tau) as more reliable markers for tracking early neuronal development. We believe that this is a promising method to track early neuronal development, particularly in the cerebellum, where microstructure changes drastically postnatally.

Results

Structural imaging confirms protracted cerebellar growth

Representative slices from intra-age non-linear registration are shown in Figure 2A, underlying the cerebellar unfolding. Both cerebellar and thalamic volumes are just below 50µl at P5 (Figure 2B). However, the segmentation at P5 is poor, as explained in Methods (Figure S1), hence data at this time point should be treated cautiously, although the cerebellum and the thalamus, which are medial line structures, are less affected by the atlas distortions.

Manganese-enhanced MRI confirms protracted cerebellar growth in rat pups.

(A) Representative slices around the thalamus and cerebellum of non-linearly averaged images acquired at P5, P15 and P30. (B) Absolute volumetric changes are shown for the segmented thalamus and cerebellum. Large dots correspond to the mean and small dots to individual data points. (C) Normalised volumetric changes. Volume at each age is normalised by the P30 volume. Large dots correspond to the mean and small dots to individual data points.

The cerebellum triples in volume, between P5 and P30, reaching 90% of its P30 volume at P20, while the thalamus reaches 90% of the P30 volume at P15.

Metabolic profile changes with age

The metabolic profile changes noticeably between P5 and P30 as shown in Figure 3A. The absolute concentrations of metabolites, estimated by LCModel using the b0 averaged spectra, shown in Figure 3A, were normalised by the water content at b0 (Figure 3B). The metabolites concentrations reported were neither corrected for T1 (TR=2.2 s) nor for the non-null diffusion-weighting and are therefore only apparent concentrations. Conventionally, metabolite absolute concentrations are normalised by the water content in the voxel or by the tCr absolute concentration, which are considered stable in the healthy brain. However, this study compares regions with different relaxation properties at different ages, with different metabolite profiles in adulthood. The water relaxation properties are different between the thalamus and cerebellum, and both water relaxation properties and content change with age34,35. The tCr concentration changes substantially with age (about +50% in the cerebrum between P7 and P28) and is known to be higher in the cerebellum in adulthood (about +40- 50% compared to the cerebrum)35,36. As an alternative to normalisation by water content, we propose in Figure S2 a normalisation by the absolute MM concentration provided by LCModel, confirming trends reported by water normalisation. MM resonances mainly come from the cytosolic proteins; they are a direct probe of the tissue and do not depend on the voxel CSF content. Although their content might slightly vary between regions in the healthy brain, they make a reasonable intracellular reference if SNR allows. Between P5 and P30, apparent metabolite concentrations increase for tNAA, Glu, Gln, and Ins, and decrease for PE and Tau (only in the thalamus). tCho and Gly exhibit slightly different trends depending on the normalisation mode (Figure S2). Note that the increase in tNAA concentration with age is protracted in the cerebellum compared to thalamus and has not reached a plateau at P30. These results generally match the metabolic concentrations measured with gold- standard classical MRS in the striatum, cortex and hippocampus of developing rats reported in Reference 35.

Metabolic profile changes postnatally.

(A) Spectra were binned per time point, region and b-value and averaged after phase and frequency correction. The SNR is clearly lower at P5/P10 in both regions: the voxel is about half of the P30 voxel size, and more animals were outliers. The SNR is also lower at P30: the rat size was slightly too big for the mouse cryoprobe at this age, so the skull is not at the top of the probe and the sensitivity is suboptimal (B) Individual spectra at b=0.035 ms/µm2 were quantified with LCModel and absolute metabolite concentrations were normalised by the water content and levelled to get Signal(tCr,P30)/Signal(water,P30)∼8 in the thalamus. See Figure S2 for a comparison with normalisation by MM.

Metabolite diffusion properties differ between regions and change with age

Thalamic (Figure 4A) and cerebellar (Figure 4B) signal attenuations up to b=30 ms/µm2 and ADCs at TM=100-1000 ms are shown for tNAA, Glu, tCr, Tau, tCho and Ins. These diffusion properties differ between both regions at each age (see Figure S3 for a region-to-region visual comparison). For most metabolites (except tCho), the signal attenuation is stronger in the thalamus, suggesting less restricted diffusion compared to the cerebellum. There is also a trend for lower signal attenuation and lower ADCs with increasing age for some metabolites, suggesting increasing restriction of metabolite diffusion. This is particularly striking for Tau in the thalamus. To give an estimation of the MRS fit performance, CRLBs are given for tNAA, Glu, tCr, Tau, tCho and Ins for the different diffusion conditions in Table S1.

Biophysical modelling issued from metabolites diffusion properties shows differential developmental trajectories in the cerebellum (A) and in the thalamus

(B). The first column (A,B) reports the signal attenuation as a function of b-value at TM=100ms, and the second column reports the ADC varying with diffusion time. Shadowed error bars represent the standard deviations. Modeling results from the spheres+”astrosticks” model is reported in the third column (A,B). P5 is shaded in the cerebellum due to possible motion artefact. The parameter space (sphere fraction, sphere radius) is displayed. Error bars represent the standard deviations. Llength, modeling output from the morphometric model is shown in the fourth column (A,B). P5 is shaded in the cerebellum due to possible motion artefact.

Biophysical modelling underlines different developmental trajectories of cell microstructure between the cerebellum and the thalamus

Data were fitted with a spheres + “astrosticks” model and a morphometric model (more details in the Methods section). The spheres + “astrosticks” model captures diffusion within spheres (representing cell bodies) and randomly oriented sticks (representing cell processes/neurites): parameters extracted are sphere fraction, fsphere (or equivalently neurite fraction as 1-fsphere) and cell radius, R. The parameter space (sphere radius vs sphere fraction) is represented in the 3rd column of Figure 4A (cerebellum) and Figure 4B (thalamus) for tNAA, Glu, tCr, Tau, tCho and Ins at all ages (see Figure S4 for a region- to-region visual comparison).

Mean values and standard deviations are reported in Table 1, along with the results from the linear mixed effect model ∼ Age + Region + Age:Region + (1|PupID). The fits on averaged data, for binned spectra by age and region, are shown in Figure S5.

The morphometric model, only applied to averaged ADCs at long td, interprets diffusion in branched structures: the segment length, Lsegment is reported on the last column of Figure 4A/B (see Figure S6 for a region-to-region visual comparison and Figure S7 for all parameters estimated by the morphometric model: Dintra, NBranch, Lsegment).

Neuronal markers: tNAA and Glu are primarily concentrated in neurons in adulthood.

Thalamus: The sphere fraction and the radius estimated from tNAA and Glu diffusion properties do not change much with age. For both metabolites, fsphere is around 0.25 and R decreases to 5-6 µm. Lsegment is stable with age for tNAA, but shows a reduction at P15-P20 for Glu.

Cerebellum: The sphere fraction and the radius estimated from tNAA diffusion properties vary with age. fsphere nears 0.5 in the cerebellum and R decreases below 5 µm for tNAA and Glu. Lsegment decreases until P15 for both tNAA and Glu.

Between regions, sphere fraction differs significantly (Table 1).

Glial markers: tCho and Ins are primarily concentrated in glial cells in adulthood. The general trend is a high sphere fraction (∼0.7-0.9) for Ins and moderately high (∼0.5-0.6) for tCho. The radius estimated from tCho differs significantly between regions. Note the trend of decreasing sphere fraction in the cerebellum with age. Data at P15 for Ins in the thalamus were individually not reliable enough to apply the spheres + astrosticks model (Table S1 and the particularly higher CRLBs at P15 for Ins).

Non-specific markers: tCr and Tau are not known to be compartmentalised in a specific cell-type.

Thalamus: Parameter estimates for tCr diffusion properties do not change substantially with age (fsphere∼0.35, R∼6.5µm). Tau parameters estimation varies a lot with age: R shows a substantial decrease and fsphere increases.

Cerebellum: fsphere decreases from 0.63 (P10) to 0.41 (P30) for tCr, but R is stable. Tau follows a similar pattern, although with an overall lower fsphere (0.43 at P10 to 0.29 at P30). Lsegment increases in the cerebellum from P10 for tCr and increases P15 for Tau.

Between regions, the sphere fraction differs significantly for both metabolites (Table 1), but only Tau exhibits a significant change with age, and this change is different between regions (significant interaction term). For Tau, the radius is significantly different between regions, changes significantly with age and this change is different between regions (significant interaction term).

Discussion

As in Reference 2 we used MEMRI to validate the protracted development of the cerebellum in rat neonates3, unfolding and tripling volume between P5 and P30. Early developmental cell-specific cerebellar and thalamic microstructural features were assessed with dMRS.

We specifically report diffusion properties for neuronal markers tNAA and Glu, glial markers tCho and Ins, and the non-specific markers tCr and Tau. Metabolites exhibit significantly different diffusion properties between regions, demonstrating the relevance of dMRS to probing early cell structure development. fsphere is notably higher in the cerebellum for metabolites compartmentalised fully (tNAA, Glu) or partially (tCr, Tau) in neurons, suggesting the influence of the cell-body dense granular layer of the cerebellum. fsphere possibly reflects the relative volume of sphere-like structures (e.g., cell bodies) compared to the relative volume of fibrous structures (e.g., cell processes, fibres).

A study of human WM using diffusion MRI reported an increase in FA and a decrease in MD with age in early life.22 In mice, the same relationship was found in both WM and GM37. We report similar dMRI trends, in both cerebellum and thalamus (Figure S8).

We also report dMRS changes with age in both regions. The general trend observed for lower signal attenuation and ADC decrease (for some metabolites) with increasing age aligns with these reported dMRI observations, and could reflect overall microstructural complexification. However, the different nature of the signal (metabolites: mostly intracellular and low compartmental exchange vs. water: ubiquitous and fast exchange) prevents close co-interpretation.

fsphere(tCr, Tau) decreases with age in the cerebellum, meaning that tCr and Tau are located in compartments where the relative contribution of sphere-like structures (e.g. cell-bodies) decreases. In other words, it reflects cerebellar cell growth and complexification (Figure 4A, Figure 5A). In the cerebellum, while soma morphologic maturation mostly happens in the first week38, Purkinje cell dendritic expansion is rapid before P15 but then continues to mature up to the end of the 4th week38,39. Tau diffusion properties might reflect the development of Purkinje cells: (i) in cats, which cannot synthesise Tau and must source it from their diet, Tau dietary deficiency delays cortical and cerebellar maturation43, meaning that Tau is also crucial in early development, and (ii) in the cerebellum, immunohistochemistry studies locate Tau in adult Purkinje cells40–42. These observations align well too with the high Tau levels we still detect in the cerebellum at P30 (Figure 3B).

The biophysical parameters are interpreted based on the cerebellar microstructural development and main results are summarised.

Italic font corresponds to possible interpretations. (A) Schematic of cerebellar neuronal development and biophysical modelling parameters interpretation. Lsegment is interpreted as the distance between two branches of the dendritic tree, NBranch as the number of embranchments. The sphere radius Rsphere is interpreted as the average soma radius (i.e. granule and Purkinje cells). fsphere,LT represents the ratio of cerebellar layer thicknesses between the layers rich in sphere-like cell components (e.g., somas in the I/EGL, GL and PL) and the layers rich in fibrous cell components (e.g., cell processes or “neurite” in the ML). fsphere,LT can be seen as a rough approximation of fsphere and is inversely related to the dendritic tree growth, see Table S2. (B) Summary of the cerebellar results coming from the modelling. (1) and (2) refer to the models depicted in Figure 1. (C) Tau seems to change compartments in the thalamus with age, going from neuronal-like compartments (low sphere fraction) to glial-like compartments (high sphere fraction), also indicated by its important signal drop with age. It suggests that Tau is a marker of neuronal maturation. Its concentration remains high up to P30 in the cerebellum. Tau properties in the cerebellum (decreasing sphere fraction + relatively low sphere fractions) and contrast with the thalamus suggest that Tau could be a marker of cerebellar neuronal development.

The diffusion properties of glial metabolites, fsphere(Ins) and fsphere(tCho), also tend to decrease with age. This could be explained by another cerebellar peculiarity: Bergmann glia develop long processes along Purkinje cells and this specific growth could contribute to the fsphere(tCr,Tau) decrease.

Validation would require an atlas of all types of cerebellar cellular processes and bodies with estimates of their relative contribution. Such a resource would be immensely valuable but does not exist. Using literature data, we use the relative thicknesses of the EGL, IGL, PL (“cell body-like” layers) and ML (“fibre-like” layer) to derive fsphere,LT, a proxy of fsphere, as illustrated in Figure 5A and reported in Table S2 and Figure S11. fsphere,LT represents the contribution of sphere-rich layers compared to fibre-rich layers. The evolution of fsphere,LT trend with age is comparable to fsphere(tCr). Although fsphere(tCr) and fsphere(Tau) have similar trends, fsphere(Tau) values are overall lower, and below fsphere,LT values (Figure S11). This lends weight to the idea that Tau is preferentially located in cerebellar neurons during cerebellar development (and potentially less so in granule cells). In comparison, tCr, which appears more evenly spread between all cerebellar cell types, tracks the evolution of fsphere,LT more closely. Note that Tau has a higher Dintra44 compared to the other metabolites, potentially affecting the parameter estimation given for Tau. As described in Methods, the fit for Tau was substantially improved for Dintra > 0.6 µm2/ms. The results for Dintra=0.7 µm2/ms are similar for all the other metabolites (Figure S9), and generally improved the estimated sphere radius standard deviation. Additionally, the evolution with age of (R,fsphere) is then very similar for tNAA and Tau, reinforcing the idea that Tau is neuronal in the cerebellum.

In the thalamus, the decrease in Tau content (Figure 3B) and the change in Tau diffusion properties (Figure 4B) point towards a change of compartmentation with age. Contrary to the trend in the cerebellum, fsphere(Tau) consistently increases with age in the thalamus (Figure S9). This means that Tau diffuses in a space with fewer and fewer processes during development, which is incompatible with the hypothesis that Tau stays in the same compartment. Neurons, but also microglia45 or astrocytes46–48 develop with increasing ramification. Neurons are already branched at P10 in the thalamus33 (shown by a low fsphere(tNAA, Glu) at P10). fsphere(Tau) at P10 nears fsphere(tNAA, Glu), indicating that Tau might be located in neuronal compartments in the neonatal thalamus. Although this hypothesis is not verifiable in existing literature, some in vitro studies show that Tau contributes to neuronal proliferation49 and neuronal growth50. At P30, fsphere(Tau) is closer to fsphere(tCho,Ins), indicating that Tau might be located in glial compartments in adulthood. Indeed, immunohistochemistry studies locate Tau in adult thalamic glia42.

We suggest that Tau should be explored further as a cerebellar marker of early neuronal development, possibly more specifically of Purkinje cells. We confirm in this study the observation from Reference 35 that both tCr and Tau have high spectroscopic signals in neonates. They are the two metabolites most reliably quantified up to P15 in both regions (continuing for Tau in the cerebellum up to P30), and the estimates of their diffusion properties are the most confident, contrary to the usual neuronal markers, tNAA and Glu. tCr stands as a good marker, too. However, there is no indication of a preferential neuronal compartmentation for this metabolite during early development, hence it is less specific to neurons, and probably even less to Purkinje cells. Because Tau seems to switch compartments in the thalamus between birth and adulthood, we do not think its diffusion properties reliably inform thalamic neuronal morphological development. In both human51,52 and rhesus monkey53 brains, Tau content increases at birth and then decreases with age, and is therefore a marker of the immature brain in both species. Although Tau concentration is low in the adult human brain, it is more concentrated in the cerebellum in adulthood52,54. A similar finding was made in the adult rhesus monkey brain53, indicating a conserved pattern across species of Tau properties in the cerebellum. These observations form a good basis for translation. Although dMRS in infants has substantial challenges, they are worth trying to overcome for a putative cerebellar marker of early neuronal development which could help understand neurodevelopmental disorders 3.

We also observed a protracted increase in tNAA signal in the cerebellum compared to the thalamus. tNAA is usually described as a neuronal osmolyte, meaning it could reflect neuronal growth and maturation and be a marker of neuronal development. However, at the voxel level, its signal reflects the neuronal volumetric contribution, i.e. differences in neuronal densities might mask the effect of development. In contrast, we want to reiterate that dMRS is a normalised measure and dMRS outcomes are not affected by the neuronal density: they represent the average morphological properties of the cell types the metabolite belongs to.

We chose to interpret the dMRS data with two biophysical models: an analytical spheres & “astrosticks” model and a computational morphometric model. The analytical model was applied simultaneously to the signal attenuation at high b-values and the ADCs at long diffusion times, and on individual animals. In addition to the valuable biophysical biomarkers, it provides a good signal representation to evaluate the significance of the differences between diffusion properties of individual datasets. The experimental design was not suitable for a randomly oriented infinite cylinders (“astrocylinders”) model as in previous studies28,29. The mixing time used for the acquisition of signal decay at high b-values (TM=100 ms) is too long to enable sensitivity to fibre radius26, hence the choice of “astrosticks” model. The sphere compartment was added to account for the high sphere fraction in the cerebellum, and for the nature of the neonatal brain (under-developed neurons). The RMSEs from the different models tested (Figure S3) tend to be more similar at P30, suggesting that it is less important to add a sphere compartment for modelling metabolites diffusion in the adult brain.

The computational morphometric model was applied to attempt to extract cell extension-related parameters, but the study design and data quality are suboptimal, and we could only reliably extract Lsegment and Dintra (Figure S7). Dintra is mostly stable with age in both regions. However, it sometimes drops or peaks, likely affecting the other parameters’ estimations. The longest mixing time (TM=1000 ms) is short compared to previous studies using a similar modelling approach27,29, meaning that the diffusion path is shorter (i.e., about 30µm in this study compared to about 45µm in previous studies, assuming Dintra=0.5 µm2/ms). Note that the higher Dintra of Tau is therefore an asset for probing early neuronal cerebellar growth. In our study, the number of embranchments (Nbranch) was very poorly evaluated because of the high level of noise in the data. However, even with a better dataset quality and a more optimal study design, it is unlikely that the Nbranch estimation could reflect the Purkinje cell morphometry because of their size and complexity in adulthood. If we had robust DTI in all animals (Figure S8), we would have included the fibre dispersion in the modelling (notably for the cerebellum), as in Reference 29. However, since we acquired enough directions for the dMRS data to be powder- averaged, we expect that the WM contribution in the cerebellum is mostly reflected by a slight fsphere drop, as exemplified by the fsphere,LT drop when including WM (Table S2).

Finally, we expected that Lsegment estimated from the morphometric model in the cerebellum would increase regularly with age, matching the dendritic tree development. For neuronal, or partly neuronal metabolites, we report an Lsegment decrease first. Except for tNAA, which consistently decreases in the cerebellum with age, Lsegment decreases until P15 for tCr, Tau and Glu and then increases, forming a U- shape curve. To better understand these results, we extracted the morphometries of developing Purkinje cells and developing thalamic interneurons (both in the mouse) from NeuroMorph.org. Interestingly, from these 3D reconstructed cells, we observe a decrease in Lsegment, 3D-recon up to P13 for Purkinje cells, whilst being quite stable in the thalamus (Figure 6), matching our non-invasive measurement trend. This could reflect the complexification of the dendritic tree following its expansion.

Morphometric parameters extracted from mouse real cells 3D reconstructions on NeuroMorph shows a decrease of Lsegment up to P13 in developing cerebellar Purkinje cells, but not in developing thalamic interneurons.

(A) Examples of cells extracted from NeuroMorph. The total number of cells used per age and region is given in Supplementary Material. References: Cerebellum P6-P1055, P10-P1356, P35-P4357; Thalamus P9-P3058 (B) The values of Branch Length (Lsegment), Branching Order and total process length are directly estimated from the morphology using the function stat_tree in the Trees Matlab Toolbox.

Conclusion

This study proposes diffusion-weighted MRS as a method to longitudinally and non-invasively measure cell-specific cerebellar and thalamic microstructural features during early development in rat neonates. We provide multiple clues indicating the sensitivity of Taurine to early neuronal cerebellar development, inviting for further validation studies and opening interest for pathological conditions.

Methods

Animals

Experiments were approved by the United Kingdom Home Office. 18 Sprague-Dawley rats (9 females) were scanned at P5, P10, P15, P20 and P30. Pups came from 3 litters (6 pups scanned per litter). Animals were induced with 4% isoflurane in oxygen and anaesthesia was maintained with 1-2% isoflurane in oxygen. Respiration was monitored with a breathing pillow. Temperature was monitored with a rectal probe, which was used as a cutaneous probe from P5 to P20 to avoid any risk of harm. Isoflurane level was adapted to keep the animals at 50-70 rpm and a warm water-circulation heating pad was used to keep the animals at 36-37°C during scanning. Pups were gently held with foam and protected tape to minimise motion, because the neonatal skull is not suitable for the use of ear bars.

For logistic reasons we could not acquire MEMRI for all litters. Animals from the second litter had manganese-enhanced anatomical scans. To this end, they were intraperitoneally injected with a 50 mg/kg solution of MnCl2 24 hours prior to the scan. Under the age of P10, the dam received the intraperitoneal injection, and the manganese was transferred to the neonates via the maternal milk2.

MRI acquisitions

Data were acquired on a 7T Biospec Bruker MRI scanner (Paravision 360.1, max gradient strength: 660 mT/m), equipped with a mouse quadrature cryoprobe, well-suited for rat neonates. An anatomical scan around the region of interest (thalamus or cerebellum) to place the voxel (T2-weighted, 2D RARE, R=8, 0.2 mm isotropic resolution, TE=33 ms, TR=3500 ms). For 6 animals, this scan was replaced by a whole brain manganese-enhanced anatomical scan (T1-weighted, 3D MGE, TE=2.4 s, TR=62 ms, 0.15 mm isotropic resolution).

dMRS acquisitions

A STELASER59 sequence (TE=34 ms, diffusion gradient duration, 8=5 ms) was used for dMRS acquisitions. The voxel was placed in the cerebellum for 9 pups, and in the thalamus for the other 9 pups. Voxel size was increased with age to follow brain growth. Diffusion-weighted spectra were acquired at 4 mixing times (TM) and 2 b-values (c.f. Table 2 for detailed parameters): a small “b0” used as a crusher for the stimulated echo, and b0+3 ms/µm2. Higher b-values were also acquired at the shortest diffusion time to probe restriction within cell bodies and enable inference of cell fraction and radius. The macromolecular baseline was acquired in the adult rodent with the same STELASER parameters (TE and 8), at the shortest TM=100 ms, but the stimulated echo was preceded by a double- inversion recovery (TI1=2000 ms, TI2=710 ms). The b-value was set to 10 ms/µm2 to reduce the residual contribution of bulk metabolites, and the tCr peak residual at 4ppm (due to the shorter T1) was corrected by subtracting a Lorentzian.

Processing

dMRS

Processing was achieved with a custom Matlab (The Mathworks) routine (https://github.com/clemoune/dmrs-neonates/tree/main/A_dMRS_processing):

Signals from different coils were combined, using the water signal at the lowest b-value and shortest diffusion time.

Outliers were identified and discarded, as described in60.

o For each transient i, the mean absolute upfield signal (range 1.9-4.5 ppm), Mup(i) was computed.

o Transients affected by motion are characterised by a clear drop of Mup(i). Transients falling below the median(Mup) - 3xMAD threshold were discarded.

o Experiments at high b-values are more susceptible to changing gradient direction (particularly if the voxel contains white matter tracts). To predict its effect across b- values and overcome the small number of acquisitions per unique condition (i.e. a (b- value, direction) pair), a diffusion kurtosis tensor was computed based on Mup. The MAD calculated at b=0 was used as a “ground truth” MAD and was reported to higher b-values (per direction). High b-values acquisitions are more easily corrupted by motion and therefore present an artificially high MAD.

Individual transients were corrected for phase and frequency drifts and averaged across all diffusion directions (powder-averaged) via an arithmetic mean. In general, metabolites signal was visually detectable on single transients (with a line-broadening of 5 Hz), except at b=20,30 ms/µm2 at P5, rendering these acquisitions more susceptible to motion artefacts.

Any water residual was removed using a singular value decomposition method centred around the water region. If the baseline was heavily distorted despite this step due to very poor water suppression, the acquisition was discarded from the analysis.

Spectra were subsequently fitted with LCModel. The simulated basis set contained the following metabolites: Ace, Ala, Asp, Cr, GABA, Gln, Glu, Gly, GPC, GSH, Ins, sIns, Lac, NAA, NAAG, PCho, PCr, Tau, PE, and included a macromolecules (MM) baseline acquired experimentally (double-inversion recovery and b=10 ms/µm2).

Diffusion properties of NAA+NAAG (tNAA), Glu, Cr+PCr (tCr), Tau, GPC+PCho (tCho) and Ins are reported. In order to model both signal compartmentalisation and restricted diffusion by cell membranes, we investigated metabolites diffusion through two sets of measurements: i) the direction-averaged signal decay as a function of the diffusion weighting b, keeping the diffusion time constant at 104.25 ms (time between diffusion gradients Δ fixed at 105.91 ms and 8 fixed at 5 ms) and varying the strength of the diffusion-sensitizing gradients; and ii) the apparent diffusion coefficient (ADC) as a function of increasing diffusion time (keeping 8 fixed at 5 ms). For the measurements i), data are reported as the ratio S(b)/S(b=0). For the measurements ii), ADCs diffusion time dependencies are reported. ADCs are calculated following . Negative ADCs were systematically removed as they point to a quantification error.

In the cerebellum, metabolites’ signal attenuation at P5 is always much stronger and quite far off the other time points. The macromolecules’ signal attenuation at high b-values is supposed to remain low59 and can be used as an internal probe for motion artefact44. However, the macromolecules’ signal attenuation is quite strong at P5 in the cerebellum (Figure S8), pointing out to a residual motion artefact despite careful processing. The spectroscopic signal in individual transients was too low at high b-values, notably in the cerebellum: it is possible that the spectral realignment failed.

MEMRI

The three first echoes of each MGE acquisition were averaged and denoised using the DenoiseImage function from the MINC toolbox, based on61. For each time point, an image with no obvious artefact and a good visual CNR was taken as the initial model and transformed into the Fischer atlas62 space. Images were then processed using registration_tamarack.py with the MAGeT segmentation in the available Pydpiper pipelines63. Due to the large deformations between P30 and P5, the P30 timepoint could not be used as a reference. An adequate atlas from P15 was generated in a couple of iterations (using a P20 atlas as intermediary). It was then used as a reference atlas and reference timepoint for the pipeline. The olfactory bulb and the more frontal parts were poorly registered because the image often had an intensity drop there, but it did not affect the registration of caudal regions, such as the cerebellum and the thalamus. Image with strong intensity differences, or corrupted by motion were removed from the analysis. The registration did not work well at P5, even when attempting to use an atlas generated at P10 (that is already slightly distorted).

Analysis

Signal decays at high-b values (at TM=100 ms) and ADCs at varying diffusion times were fitted together with a spheres and “astrosticks” (corresponding to randomly oriented sticks) model using the functions implemented in DIVE (https://git.fmrib.ox.ac.uk/fsl/DIVE). The DIVE toolbox uses the Gaussian phase distribution approximation We fitted the signal to: Free parameters were R (sphere radius) and fsphere (sphere fraction). Dintra (i.e. the metabolite intracellular free diffusivity) was fixed for both compartments at 0.5 µm2/ms. Previous studies measuring brain metabolites diffusion properties at ultra-short diffusion-times44,64 estimated the intracellular free diffusivity at 0.5-0.6 µm2/ms for most metabolites (except Tau). However other studies using different (less direct) approaches reported lower Dintra27,28, hence we tested a range of reasonable Dintra. Dintra was fixed 0.3, 0.4, 0.5, 0.6, 0.7 and 0.8 µm2/ms. RMSEs between model fit and data were equivalent for all Dintra, except for Tau in the thalamus where increasing Dintra improved the fit, particularly until Dintra=0.5µm2/ms. Moreover, parameters estimation was unrealistic for low Dintra. Increasing Dintra between 0.5 µm2/ms and 0.7 µm2/ms did not change the trend of the results, except for Tau in the thalamus (see Figure S0 and Discussion). The fit was made robust to missing data (when excluded during processing).

To test for the model accounting best for our data, we tested different analytical models. RMSE between model fit and data were calculated for “astrosticks” only (1 free parameter, Dintra), “astrocylinders” (corresponding to randomly oriented infinite cylinders, 2 free parameters: Dintra and cylinder radius), spheres and astrosticks without constraint (3 free parameters) and spheres and astrosticks with Dintra fixed at 0.5 µm2/ms (Figure S0). RMSE were the lowest for spheres and astrosticks, and were very similar with and without the Dintra constraint. Hence the model with the least free parameters was chosen. Moreover, the results from the constrained spheres and astrosticks model were more stable and physically meaningful with the constraint on Dintra. The code used for the analysis can be found on: https://github.com/clemoune/dmrs-neonates/tree/main/B_Astrosticks_Spheres_model.

Long diffusion times:

Average ADCs at varying diffusion times were fitted with the morphometric model introduced in27. Modeling was performed using the dictionary generated for29 with TM = 100, 750 ms predicted by interpolation, and a random forest regressor (‘treeBagger’ function in MatlabR2020b, with tree depth = 20; number of trees = 200 and other model hyper-parameters as per default) trained in a fully supervised fashion, using the ADCs time dependencies as input and the ground truth morphological parameters as targets; as loss function the mean squared error between predicted and ground truth parameters values was used. Parameters fitted were Dintra, Nbranch, Lsegment, SDNbranch and SDLsegment. Boundaries for fitting the model to data were respectively [0.1, 1], [2, 25], [5, 100], [2, 3] and [5, 10]. Given the data noise, only Lsegment and Dintra are estimated robustly. Fits for all parameters are shown in Supplementary Information.

Statistical analyses

Statistical analyses were run with R, using the “lmerTest65 package. To assess whether diffusion properties were significantly different in the 2 regions during development, we applied a linear mixed effect model to the parameters extracted from the spheres + ”astrosticks” model (sphere fraction and sphere radius), using age and region as fixed effects, and a random intercept for pups ID. An analysis of variance enabled to identify whether the inclusion of an interaction term age*region was necessary or not. Results were corrected for multiple hypotheses testing with a Bonferroni correction (6 metabolites,2 parameters = 12 hypotheses). The threshold corrected p-value is 4.2e-3 for α=0.05 and 8.3e-4 for α=0.01.

Funding

This work was funded by Wellcome. The Wellcome Centre for Integrative Neuroimaging is supported by core funding from the Wellcome Trust (203139/Z/16/Z and 203139/A/16/Z). For the purpose of Open Access, the corresponding author has applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission. MP is supported by UKRI Future Leaders Fellowship (MR/T020296/2).